```{r data generation, echo = FALSE, results = "hide"}

# Joint Probability Tables
x_vals <- sort(sample(1:15, 3))
x_cuts <- sort(round(runif(2, 0.01, 0.99), 2))

p1 <- x_cuts[1]
p2 <- x_cuts[2] - x_cuts[1]
p3 <- 1 - (p1 + p2)

pmf_X <- function(x){
    ifelse(x == x_vals[1], p1, ifelse(x == x_vals[2], p2, p3))
}

pmfX <- paste0("$$p_{X}(x) = \\begin{cases}", p1, " & x = ", x_vals[1], "\\\\", p2, " & x = ", x_vals[2], "\\\\", p3, " & x = ", x_vals[3], " \\end{cases}.$$")

y_vals <- sort(sample(1:15, 3))

format <- sample(1:3, 1) 

if(format == 1) {
    # P(Y|X) = xy/k
    K <- sum(y_vals * rep(x_vals, each = 3))

    pmf <- paste0("$$p_{Y|X}(y|x) = \\dfrac{xy}{",K,"}, \\quad y \\in \\{", y_vals[1], ", ", y_vals[2], ", ", y_vals[3], "\\};  x \\in \\{", x_vals[1], ", ", x_vals[2], ", ", x_vals[3], "\\}.$$")
    pmf_YX <- function(y, x) {
        x*y/K
    }
} else if(format == 2) {
    # P(Y|X) = (x+y)/k
    K <- sum(rep(x_vals, each = 3) + y_vals)

    pmf <- paste0("$$p_{Y|X}(y|x) = \\dfrac{x+y}{",K,"}, \\quad y \\in \\{", y_vals[1], ", ", y_vals[2], ", ", y_vals[3], "\\};  x \\in \\{", x_vals[1], ", ", x_vals[2], ", ", x_vals[3], "\\}.$$")
    pmf_YX <- function(y, x) {
        (x + y)/K
    }
} else {
    # P(Y|X) = y^x / k 

    K <- sum((y_vals - rep(x_vals, each = 3))^2)

    pmf <- paste0("$$p_{Y|X}(y|x) = \\dfrac{(y - x)^2}{",K,"}, \\quad y \\in \\{", y_vals[1], ", ", y_vals[2], ", ", y_vals[3], "\\};  x \\in \\{", x_vals[1], ", ", x_vals[2], ", ", x_vals[3], "\\}.$$")
    pmf_YX <- function(y, x) {
        (y - x)^2/K
    }

}

Xa <- sample(x_vals, 1)

ans1 <- sum(pmf_YX(y_vals, Xa) * pmf_X(Xa) * y_vals)
ans2 <- sum(sapply(x_vals, FUN = function(x) sum(pmf_YX(y_vals, x) * pmf_X(x) * y_vals)))

options_A <- c("E[Y|X]", "E[X|Y]", "E[X+Y|X]", "X", "Y", "X+Y")
options_B <- c("E[X]", "E[Y]", paste0("E[X|Y=", sample(y_vals, 1), "]"), paste0("E[Y|X=", sample(x_vals,1), "]", "E[X+Y]"))
type <- sample(c("a random quantity", "not a random quantity"), 1)
ans3 <- c(TRUE, rep(FALSE, 4))

if(type == "random") {
    qtext3 <- c(sample(options_A, 1), sample(options_B, 4))
} else {
    qtext3 <- c(sample(options_B, 1), sample(options_A, 4))
}
order <- sample(1:5)

ans3 <- ans3[order]
qtext3 <- qtext3[order]

id <- sprintf("%010d", sample(0:1000000000, 1))
```


Question
========

The conditional distribution of $Y$ given $X$ has a probability mass function `r pmf`

Moreover, the probability mass function for $X$ is `r pmfX`

a. What is $E[Y|X = `r Xa`]$? ##ANSWER1##
b. What is $E[Y]$? ##ANSWER2##
c. Which of the following is `r type`? 

##ANSWER3##

Question ID: *`r id`*


```{r questionlist, echo = FALSE, results = "asis"}
answerlist(qtext3, markup = "markdown")
```


Meta-information
================
extype: cloze
exclozetype: num|num|schoice
exsolution: `r fmt(ans1, 3)`|`r fmt(ans2, 3)`|`r mchoice2string(ans3)`
exname: ConditionalExpectedValue