```{r data generation, echo = FALSE, results = "hide"}

sided <- sample(1:3, 1)
n <- sample(4:8, 1)
possible_rows <- expand.grid(rep(list(c(-1, 1)), n)) |> as.matrix()
ranking <- possible_rows[sample(1:nrow(possible_rows), 1), ]

## For each scenario; specify mu and SD 
sc <- sample(1:3, 1)
if(sc == 1) {
    mu <- sample(500:800, 1)
    sd <- sample(70:80, 1)
    rd <- 2 

    qscenario <- "A local environmental group is interested in understanding the median daily water consumption, per household, within their city."

    directions <- c(
        paste0("They claim that the median usage exceeds ", mu, " liters."),
        paste0("They claim that the median usage is less than ", mu, " liters."),
        paste0("They claim that the median usage differs from the city-reported median of ", mu, " liters.")
    )
} else if(sc == 2) {
    mu <- sample(40:100, 1)
    sd <- sample(5:15, 1)
    rd <- 0 

    qscenario <- "A sports coach is interested in understanding the number of push-ups that their athletes can complete."
    directions <- c(
        paste0("They suspect that the median number of push-ups exceeds ", mu, "."),
        paste0("They suspect that the median number of push-ups is less than ", mu, "."),
        paste0("They suspect that the median number of push-ups is not ", mu, ".")
    )
} else if(sc == 3) {
    mu <- 0
    sd <- sample(1:5, 1)
    rd <- 0

    qscenario <- "A therapist wants to investigate the impact of a new mindfulness technique on their patients' stress scores. They measure the change in stress score (on a standardized scale) that was observed after a single week of implementing the new practice."

    directions <- c(
        paste0("They wish to determine whether the practice results in an increase in stress levels."),
        paste0("They wish to determine whether the practice results in a decrease in stress levels. "),
        paste0("They wish to determine whether the practice results in a change in stress levels.")
    )

}

## Generate unique values for 'X' 
X <- round(unname(sort(abs(rnorm(n, 0, sd)))), rd)
if(any(X == 0)) {
    X[which(X==0)] <- 1/(10^rd)
}

if(length(unique(X)) < n) {
    # There are duplicates
    for(idx in 1:(n-1)) {
        idxs <- idx + which(X[(idx+1):n] == X[idx])
        X[idxs] <- X[idx] + 1/(10^rd)
    }
}

X <- X * ranking + mu
U <- sum(ranking * 1:n)

X <- X[sample(1:n, n)]

# Correct the obvious concerns
# Sided = 1: X > Y
# Sided = 2: X < Y
# Sided = 3: X = Y
if((sided == 1 && U < 0) || (sided == 2 && U > 0)) {
    sided <- 3 
}

direction <- directions[sided]

# Draw conclusions:
## Specify alpha based on p-value and desired conclusion.
## If p-value is greater than 0.2, change conclusion to 'fail to reject'
## If p-value is less than 0.01, change conclusion to 'reject'

if(sided == 1) {
    p_val <- 1 - psignrank((U-1)/2 + n*(n+1)/4, n = n)
} else if(sided == 2) {
    p_val <- psignrank(-U/2 + n*(n+1)/4, n = n)
} else {
    p_val <- 2 * psignrank(-1*abs(U)/2 + n*(n+1)/4, n = n)
}

if(p_val >= 0.2) {
    conclusion <- 1
} else if(p_val <= 0.01) {
    conclusion <- 0
} else {
    conclusion <- sample(0:1, 1)
}

if(conclusion == 0) {
    # Reject H0
    alpha <- round(runif(1, p_val + 0.01, 0.2), 2)
} else {
    # Fail to Reject
    alpha <- round(runif(1, 0.01, min(p_val + 0.01, 0.2)), 2)
}

incorrect_conclusions <- c(
    "Because the $p$-value is greater than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we fail to reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we accept the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we accept the null hypothesis."
)
correct_conclusion <- ifelse(p_val <= alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)
incorrect_conclusion_include <- ifelse(p_val > alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)

answer_c <- c(TRUE, rep(FALSE, 4))
options_c <- c(correct_conclusion, incorrect_conclusion_include, sample(incorrect_conclusions, 3))
sort_c <- sample(1:5, 5)

ans1 <- U
ans2 <- p_val
ans3 <- answer_c[sort_c]

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========

`r qscenario`

`r direction`

They observe the following data: 

> `r paste0(X, collapse=", ")`.

a. What is the value of the test statistic used to test the given hypothesis. Use average rank tie-breaking, if required. ##ANSWER1##
b. What is the $p$-value computed for the hypothesis test? ##ANSWER2##
c. What is the conclusion from the hypothesis test, at a significance level of `r alpha`.

##ANSWER3##

Question ID: *`r id`*

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(options_c[sort_c], markup = "markdown")
```


Meta-information
================
extype: cloze
exsolution: `r fmt(ans1, 3)`|`r fmt(ans2, 3)`|`r mchoice2string(ans3)`
exclozetype: num|num|schoice
exname: TwoSampleZTest
extol: 0.01