```{r data generation, echo = FALSE, results = "hide"}

N <- sample(8:16, 1)
n <- sample(3:(N-3), 1)
m <- N - n 

ranking <- rep("Y", N)
ranking[sample(1:N, n, FALSE)] <- "X"

## For each scenario; specify mu and SD 
sc <- sample(1:3, 1)
if(sc == 1) {
    mu <- 6
    sd <- 1
    rd <- 2 

    qscenario <- "A botanist wishes to compare the growth of two different varieties of tomato plants. The heights of plants from two different varieties are recorded, after two weeks of growing. The botanist is interested in understanding whether the growth patterns differ for these varieties, or not."
} else if(sc == 2) {
    mu <- 70
    sd <- 8
    rd <- 0

    qscenario <- "A teacher wants to determine if there is a significant difference in the exam scores of students who attended a review session and those who did not. They collect the exam scores for individuals who attended the review, and for those who did not. Because the students who attended the review may have been substantially different from those who did not, they are simply interested in understanding whether there is any difference in the scores."
} else if(sc == 3) {
    mu <- round(runif(1, 200, 800), 0)
    sd <- round(runif(1, 20, 100), 0)
    rd <- 0

    qscenario <- "A researcher wants to compare the reaction times of two groups of participants: one group that has been drinking water and another group that has been drinking caffeine. Past research suggests that the distribution of reaction times in the population may change in location under different conditions, but it does not change in shape. They want to test whether caffeine changes the median of the distribution."
}

## Generate unique values for 'X' 
combined_data <- round(unname(sort(rnorm(N, mu, sd))), rd)
if(any(combined_data == 0)) {
    combined_data[which(combined_data==0)] <- 1/(10^rd)
}

if(length(unique(combined_data)) < N) {
    # There are duplicates
    for(idx in 1:(N-1)) {
        idxs <- idx + which(combined_data[(idx+1):N] == combined_data[idx])
        combined_data[idxs] <- combined_data[idx] + 1/(10^rd)
    }
}

X <- combined_data[which(ranking == "X")]
Y <- combined_data[which(ranking == "Y")]

Rank_X <- rank(combined_data)[which(combined_data %in% X)]
Rank_Y <- rank(combined_data)[which(combined_data %in% Y)]

Ux <- n*m + n*(n+1)/2 - sum(Rank_X)
Uy <- n*m + m*(m+1)/2 - sum(Rank_Y)

U <- min(Ux, Uy)

X <- X[sample(1:n, n)]
Y <- Y[sample(1:m, m)]

# Draw conclusions:
## Specify alpha based on p-value and desired conclusion.
## If p-value is greater than 0.2, change conclusion to 'fail to reject'
## If p-value is less than 0.01, change conclusion to 'reject'

p_val <- 2 * min(pwilcox(U, n=n, m=m), 1-pwilcox(U-1, n=n, m=m))


if(p_val >= 0.2) {
    conclusion <- 1
} else if(p_val <= 0.01) {
    conclusion <- 0
} else {
    conclusion <- sample(0:1, 1)
}

if(conclusion == 0) {
    # Reject H0
    alpha <- round(runif(1, p_val + 0.01, 0.2), 2)
} else {
    # Fail to Reject
    alpha <- round(runif(1, 0.01, min(p_val + 0.01, 0.2)), 2)
}

incorrect_conclusions <- c(
    "Because the $p$-value is greater than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we fail to reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we accept the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we accept the null hypothesis."
)
correct_conclusion <- ifelse(p_val <= alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)
incorrect_conclusion_include <- ifelse(p_val > alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)

answer_c <- c(TRUE, rep(FALSE, 4))
options_c <- c(correct_conclusion, incorrect_conclusion_include, sample(incorrect_conclusions, 3))
sort_c <- sample(1:5, 5)

ans1 <- U
ans2 <- p_val
ans3 <- answer_c[sort_c]

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========

`r qscenario`

They observe the following data: 

> $X$: `r paste0(X, collapse=", ")`.

> $Y$: `r paste0(Y, collapse=", ")`.

a. What is the value of the test statistic used to test the given hypothesis. Use average rank tie-breaking, if required. ##ANSWER1##
b. What is the $p$-value computed for the hypothesis test? ##ANSWER2##
c. What is the conclusion from the hypothesis test, at a significance level of `r alpha`.

##ANSWER3##

Question ID: *`r id`*

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(options_c[sort_c], markup = "markdown")
```


Meta-information
================
extype: cloze
exsolution: `r fmt(ans1, 3)`|`r fmt(ans2, 3)`|`r mchoice2string(ans3)`
exclozetype: num|num|schoice
exname: TwoSampleZTest
extol: 0.01