```{r data generation, echo = FALSE, results = "hide"}

sided <- sample(1:3, 1)
n <- sample(4:8, 1)

ranks <- sample(1:(n-1), n, TRUE)

possible_rows <- expand.grid(rep(list(c(-1, 1)), n)) |> as.matrix()
signs <- possible_rows[sample(1:nrow(possible_rows), 1), ]

## For each scenario; specify mu and SD 
sc <- sample(1:3, 1)
if(sc == 1) {
    mu <- 0
    sd <- 5
    meanX <- 25
    sdX <- 3

    rd <- 0 

    qscenario <- "A therapist wants to assess the impact of a new treatment for depression, after a single session of its use. They record the depressive scores (on a standardized scale) for their patients before ($Y$) and after ($X$) therapy."

    directions <- c(
        "They hypothesize that a single session of the new method will increase depressive scores.",
        "They hypothesize that a single session of the new method will decrease depressive scores.",
        "They question whether a single sesion of the new method will change depressive scores."
    )
} else if(sc == 2) {
    mu <- 0
    sd <- 3
    meanX <- 140
    sdX <- 5

    rd <- 0 

    qscenario <- "A doctor wants to assess the impact that a new medication will have on patient systolic blood pressure. They take a group of individuals and measure their blood pressures both before ($Y$) and after ($X$) starting the medication."

    directions <- c(
        "They hypothesize that the medication will increase systolic blood pressure.",
        "They hypothesize that the medication will lower systolic blood pressure.",
        "They question whether the medication will change systolic blood pressure."
    )
} else if(sc == 3) {
    mu <- 0
    sd <- 5
    meanX <- 25
    sdX <- 5

    rd <- 0 

    qscenario <- "A researcher wants to determine the impact of an administered substance on the quality of sleep that individuals' receive. Using a standardized sleep scoring system, the researchers measure the quality of sleep for a group of individuals both after ($X$) and before ($Y$) the substance is administered."

    directions <- c(
        "The researchers suspect that the substance will increase the reported sleep quality.",
        "The researchers suspect that the substance will decrease the reported sleep quality.",
        "The researchers question whether the substance will have any impact on the reported sleep quality."
    )

}

## Generate unique values for 'X' 
D <- round(unname(sort(abs(rnorm(length(unique(ranks)), 0, sd)))), rd)


if(length(unique(D)) < length(unique(ranks))) {
    # There are duplicates
    for(idx in 1:(length(unique(ranks))-1)) {
        idxs <- idx + which(D[(idx+1):length(unique(ranks))] == D[idx])
        D[idxs] <- D[idx] + 1/(10^rd)
    }
}

D <- rep(D, times=as.numeric(table(ranks)))
D <- unname(D * signs + mu)
U <- sum(rank(abs(D-mu)) * signs * (X != mu))


X <- round(rnorm(n, mean = meanX, sd = sdX), rd)
Y <- X - D

order <- sample(1:n, n)

X <- X[order]
Y <- Y[order]


# Correct the obvious concerns
# Sided = 1: X > Y
# Sided = 2: X < Y
# Sided = 3: X = Y
if((sided == 1 && U < 0) || (sided == 2 && U > 0)) {
    sided <- 3 
}

direction <- directions[sided]

# Draw conclusions:
## Specify alpha based on p-value and desired conclusion.
## If p-value is greater than 0.2, change conclusion to 'fail to reject'
## If p-value is less than 0.01, change conclusion to 'reject'

if(sided == 1) {
    p_val <- 1 - pnorm(U/sqrt(n*(n+1)*(2*n+1)/6))
} else if(sided == 2) {
    p_val <- pnorm(U/sqrt(n*(n+1)*(2*n+1)/6))
} else {
    p_val <- 2 * pnorm(-1*abs(U)/sqrt(n*(n+1)*(2*n+1)/6))
}

if(p_val >= 0.2) {
    conclusion <- 1
} else if(p_val <= 0.01) {
    conclusion <- 0
} else {
    conclusion <- sample(0:1, 1)
}

if(conclusion == 0) {
    # Reject H0
    alpha <- round(runif(1, p_val + 0.01, 0.2), 2)
} else {
    # Fail to Reject
    alpha <- round(runif(1, 0.01, min(p_val + 0.01, 0.2)), 2)
}

incorrect_conclusions <- c(
    "Because the $p$-value is greater than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we fail to reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we accept the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we accept the null hypothesis."
)
correct_conclusion <- ifelse(p_val <= alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)
incorrect_conclusion_include <- ifelse(p_val > alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)

answer_c <- c(TRUE, rep(FALSE, 4))
options_c <- c(correct_conclusion, incorrect_conclusion_include, sample(incorrect_conclusions, 3))
sort_c <- sample(1:5, 5)

ans1 <- U
ans2 <- p_val
ans3 <- answer_c[sort_c]

data_string <- paste0("* Observation ", 1:n, ": X=", X, " and Y=", Y, ".", collapse = "\n")

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========

`r qscenario`

`r direction`

They observe the following data: 

`r I(data_string)`

a. What is the value of the test statistic used to test the given hypothesis. Use average rank tie-breaking, if required. ##ANSWER1##
b. What is the $p$-value computed for the hypothesis test? ##ANSWER2##
c. What is the conclusion from the hypothesis test, at a significance level of `r alpha`.

##ANSWER3##

Question ID: *`r id`*

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(options_c[sort_c], markup = "markdown")
```


Meta-information
================
extype: cloze
exsolution: `r fmt(ans1, 3)`|`r fmt(ans2, 3)`|`r mchoice2string(ans3)`
exclozetype: num|num|schoice
exname: TwoSampleZTest
extol: 0.01