```{r data generation, echo = FALSE, results = "hide"}

scenarios <- list(
    list(
        scenario = "In a survey conducted among students, they were asked about their favourite colors. The options were red, blue, green, yellow, and orange.",
        options = c("red", "blue", "green", "yellow", "orange")
    ),
    list(
        scenario = "A group of students was surveyed about their preferred leisure activities. The options included reading, playing sports, watching movies, listening to music, and playing video games.",
        options = c("reading", "playing sports", "watching movies", "listening to music", "playing video games")
    ),
    list(
        scenario = "A group of students was asked for their favourite genre of music from rock, pop, hip-hop, jazz, and classical.",
        options = c("rock", "pop", "hip-hop", "jazz", "classical")
    ),
    list(
        scenario = "A survey was conducted to find out how many pets each household owns.",
        options = c("0", "1", "2", "3", "4+")
    ),
    list(
        scenario = "The number of books read by students over the summer break was collected.",
        options = c("0", "1", "2", "3", "4+")
    )
)

n <- sample(8:15, 1)
sc_idx <- sample(1:length(scenarios), 1)
scenario <- scenarios[[sc_idx]]$scenario
options <- scenarios[[sc_idx]]$options

observations <- sample(options, n, TRUE)

table_output <- data.frame(
    Option = options,
    Frequency = c("(a)", "(b)", "(c)", "(d)", "(e)"),
    'Relative Frequency' = c("(f)", "(g)", "(h)", "(i)", "(j)"),
    check.names=FALSE
)



as <- sapply(options, function(x) {sum(observations == x)})
a1 <- as[1]
a2 <- as[2]
a3 <- as[3]
a4 <- as[4]
a5 <- as[5]

as_r <- as/sum(as)
a6 <- as_r[1]
a7 <- as_r[2]
a8 <- as_r[3]
a9 <- as_r[4]
a10 <- as_r[5]

tbl_str <- knitr::kable(table_output, format = "markdown", align='c')

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========
`r scenario`

Below are `r n` responses. 

> `r observations`

Using these data, consider the following frequency distribution. 

```{r freq_dist, echo = FALSE, results = "asis"}
tbl_str
```
Fill in each of the corresponding blanks.

(a) ##ANSWER1##
(b) ##ANSWER2##
(c) ##ANSWER3##
(d) ##ANSWER4##
(e) ##ANSWER5##
(f) ##ANSWER6##
(g) ##ANSWER7##
(h) ##ANSWER8##
(i) ##ANSWER9##
(j) ##ANSWER10##

(Question ID: *`r id`*)

Meta-information
================
exname: Frequency Distribution (Categorical and Discrete)
extype: cloze
exclozetype: num|num|num|num|num|num|num|num|num|num
exsolution: `r fmt(a1, 0)`|`r fmt(a2, 0)`|`r fmt(a3, 0)`|`r fmt(a4, 0)`|`r fmt(a5, 0)`|`r fmt(a6, 3)`|`r fmt(a7, 3)`|`r fmt(a8, 3)`|`r fmt(a9, 3)`|`r fmt(a10, 3)`
extol: 0|0|0|0|0|0.01|0.01|0.01|0.01|0.01