```{r data generation, echo = FALSE, results = "hide"}

generate_histogram_data <- function(n = 1000, modality, symmetry, skewness) {
  if (!modality %in% c(1, 2, 3)) {
    stop("Modality must be 1 (unimodal), 2 (bimodal), or 3 (multimodal).")
  }
  if (!symmetry %in% c(1, 2)) {
    stop("Symmetry must be 1 (symmetric) or 2 (asymmetric).")
  }
  if (!skewness %in% c(1, 2, 3)) {
    stop("Skewness must be 1 (right), 2 (left), or 3 (none, but asymmetric).")
  }

  if (modality == 1) {
    # Unimodal
    if (symmetry == 1) {
      # Symmetric (Normal distribution)
      data <- rnorm(n)
    } else {
      # Asymmetric
      if (skewness == 1) {
        # Right-skewed (Exponential or Gamma-like)
        data <- rexp(n, rate = 1)
      } else if (skewness == 2) {
        # Left-skewed (Reflected Gamma-like)
        data <- -rexp(n, rate = 1)
      } else {
        # Asymmetric, no specific skew (e.g., a shifted distribution)
        data <- c(runif(round(0.9 * n, 0), 0, 30), rnorm(round(0.1 * n, 0), sample(c(5:10, 20:25), 1), 1))
        
      }
    }
  } else if (modality == 2) {
    # Bimodal (Mixture of two normals)
    if (symmetry == 1) {
      # Symmetric bimodal (Means equidistant from 0)
      mean1 <- sample(-10:-2, 1)
      mean2 <- -1 * mean1
      group <- rbinom(n, 1, 0.5)
      data <- ifelse(group == 0, rnorm(n, mean1, 1), rnorm(n, mean2, 1))
    } else {
      # Asymmetric bimodal (Means not equidistant or different SDs)
        n1 <- round(runif(1, 0.25, 0.35) * n, 0)
        n2 <- n - n1

        if(sample(0:1, 1) == 1) {
            n1 <- n - n1
            n2 <- n - n2
        }

        mean1 <- sample(-10:-2, 1)
        mean2 <- sample(max(2, mean1+5):10, 1)

        d1 <- rnorm(n1, 0, 1) + mean1
        d2 <- rnorm(n2, 0, 1) + mean2 

        if (skewness == 1) {
            m <- sum(d2 > mean2)
            d3 <- rexp(m, rate = 1/mean2)
            d2 <- c(d2[which(d2 < mean2)], d3[which(d3 > mean2)])
        } else if (skewness == 2) {
            m <- sum(d1 < mean1)
            d3 <- -1*rexp(m, rate = 1/abs(mean1))
            d1 <- c(d1[which(d1 > mean1)], d3[which(d3 < mean1)])
        }

        data <- c(d1, d2)

        data <- data[which(data > 3 * mean1 & data < 3 * mean2)]
    }
  } else {
    # Multimodal (Mixture of multiple normals)

    if(symmetry == 1) {
        # Take 50% at middle and then 25% at lower models 
        mean1 <- sample(-10:-3, 1)
        mean2 <- -1 * mean1
        group <- rbinom(n, 1, 0.5)
        data <- ifelse(group == 0, rnorm(n, mean1, 1), rnorm(n, mean2, 1))
        data <- c(data, rnorm(n, 0, 1))
    } else {
        # Asymmetric bimodal (Means not equidistant or different SDs)
        
        mean1 <- sample(-10:-2, 1)
        mean2 <- sample(max(2, mean1+5):10, 1)
        mean3 <- mean1 - 5
        mean4 <- mean2 + 5 

        n1 <- round(runif(1, 0.35, 0.45) * n / 2, 0)
        n2 <- n - n1

        if(sample(0:1, 1) == 1) {
            n1 <- n - n1
            n2 <- n - n2
        }

        d1 <- rnorm(n1, 0, 1) + mean1
        d2 <- rnorm(n2, 0, 1) + mean2 
        d3 <- rnorm(n2, 0, 1) + mean3
        d4 <- rnorm(n1, 0, 1) + mean4

        if (skewness == 1) {
            m <- sum(d4 > mean4)
            d5 <- rexp(m, rate = 1/mean4)
            d4 <- c(d4[which(d4 < mean4)], d5[which(d5 > mean4)])
        } else if (skewness == 2) {
            m <- sum(d3 < mean3)
            d5 <- -1*rexp(m, rate = 1/abs(mean3))
            d3 <- c(d3[which(d3 > mean3)], d5[which(d5 < mean3)])
        }

        data <- c(d1, d2, d3, d4)

        data <- data[which(data > 3 * mean3 & data < 3 * mean4)]
    }
  }
  return(data)
}

n <- 100000

modality <- sample(1:3, 1)
symmetry <- sample(1:2, 1)
skewness <- sample(1:3, 1)

if(symmetry == 1) {
    skewness <- 3
}

data <- generate_histogram_data(n, modality, symmetry, skewness)

ans1 <- c(FALSE, FALSE, FALSE)
ans2 <- c(FALSE, FALSE)
ans3 <- c(FALSE, FALSE, FALSE)

ans1[modality] <- TRUE
ans2[symmetry] <- TRUE
ans3[skewness] <- TRUE

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========
Consider the following histogram.

```{r histogram, warning = FALSE, echo = FALSE, results = "hide", fig.height = 5, fig.width = 6, fig.path = "", fig.cap = ""}
library(ggplot2)
ggplot(data.frame(x = data), aes(x=x)) + geom_histogram(binwidth = 1) + labs(y = "Frequency")
```

a. Describe the modality of the histogram. ##ANSWER1##
b. Describe the symmetry of the histogram. ##ANSWER2##
c. Describe the skewness of the histogram. ##ANSWER3##

(Question ID: *`r id`*)

Answerlist
----------
* Unimodal
* Bimodal
* Multimodal
* Symmetric
* Asymmetric
* Right-Skew
* Left-Skew 
* No Skew

Meta-information
================
exname: Histogram Shape
extype: cloze
exclozetype: schoice|schoice|schoice
exsolution: `r mchoice2string(ans1, single=TRUE)`|`r mchoice2string(ans2, single=TRUE)`|`r mchoice2string(ans3, single=TRUE)`