# Introduction

## What is Probability?
At its core, statistics is the study of uncertainty. Uncertainty permeates the world around us, and in order to make sense of the world, we need to make sense of uncertainty. The language of uncertainty is **probability**. Probability is a concept which we all likely have some intuitive sense of. If there was a 90% probability of rain today, you likely considered grabbing an umbrella. You are not likely to wager your life savings on a game that has only a 1% probability of paying out. We have a sense that probability provides a measure of *likelihood*. Defining probability mathematically is a non-trivial task, and there have been many attempts to formalize it throughout history. While we will spend a good deal of time formalizing notions of probability in this course, we first pause to emphasize the familiarity with probability that you are likely starting with.

Suppose that two friends, Charles and Sadie, meet for coffee once a week. During their meetings they have wide-ranging, deep, philosophical conversations spanning across many important topics.^[For instance they ask "do we all really see green as the same colour?" or "why is it that 'q' comes as early in the alphabet as it does? it deserves to be with 'UVWXYZ'?"] Beyond making progress on some of the most pressing issues of our time, Charles and Sadie each adore probability. As a result, at the end of each of their meetings, they play a game to decide who will pay. The game proceeds by having them flip a coin three times. If two or more heads come up Charles pays, and otherwise Sadie pays.

We can think about the strategy that they are using here and *feel* that this is going to be "fair". With two or more heads, Charles pays. With two or more tails, Sadie pays. There always has to be either two or more heads *or* two or more tails, and each is equally likely to come up^[There has been some recent literature, see @fairCoin, which may suggest that a coin is ever so slightly more likely to land on the same side it started on, perhaps undermining this assertion.]. The outcome of their game is uncertain before it begins, but we know that in the long run neither of the friends is going to be disadvantaged relative to the other. We can say that the probability that either of them pays is equal. It's 50-50. Everything is balanced.

Now imagine that one day, in the middle of their game, Charles gets a very important phone call ^[Someone has just pointed out the irony in the fact that there is no synonym for synonym. Technically, there are the words *metonym* and *poecilonym* and *polyonym*, but these are rarely used and Charles would wager that there is a very high probability you have never seen them.] and he leaves abruptly after the first coin has been tossed. The first coin toss showed a heads. Sadie, recognizing the gravity of the phone call, pays for the both of them, but she also realizes that Charles was well on the way to having to pay.

:::{#exm-basic-prob}
## Basic Probability Enumeration

What is the probability that Sadie would have had to pay in the aforementioned scenario? That is, assuming that the first coin shows a head, what is the probability that at least two heads are shown on the first three coin tosses?

::::{.callout .solution collapse="true"}
## Solution
Sadie figures that any coin toss is equally likely to show heads or tails. because the first coin showed heads, then there are four possible sequences that could have shown up:

1. $H,H,H$;
2. $H,H,T$;
3. $H,T,H$;
4. $H,T,T$.

In three of these situations ((1) $H,H,H$, (2) $H,H,T$, and (3) $H,T,H$) there are two heads and so Charles would have to pay. In one of them there are two tails, and so Sadie would have to pay. As a result, Charles would have to pay in $3$ of the $4$ (with probability $0.75$) and Sadie in $1$ of the $4$ (with probability $0.25$).
::::
:::

Looking at @exm-basic-prob, we can see that Sadie should likely have not paid. Only one out of every four times would she have had to pay, given the first coin being heads. However, we can not be certain that, had all three tosses been observed, Sadie would *not* have paid. It is possible that we would have observed two tails, making her responsible for the bill. This possibility happens one time out of four, which is more likely than the probability of rolling a four on a six-sided die^[This event happens one time out of every six.]. Fours are rolled on six-sided dice quite frequently^[Again, about one out of every six rolls], and so it is not all together unreasonable for her to have paid.

This seemingly simple concept is the core of probability. Probability serves as a method for quantifying uncertainty. It allows us to make numeric statements regarding the set of outcomes that we can observe, by quantifying the frequency with which we expect to observe those outcomes. Probability does not *remove* the uncertainty. We still need to flip the coin or roll the die to know what will happen. All probability gives us is a set of tools to quantify this uncertainty. These tools are critical for decision making in the face of the ever-present uncertainty around us.

## How to Interpret Probabilities (like a Frequentist)
We indicated that, intuitively, probability is a measure of the frequency with which a particular outcome occurs. This intuition can be codified exactly with the **Frequentist interpretation** of probability. According to the Frequentist interpretation (or frequentism, as it is often called), probabilities correspond to the proportion of time any event of interest is^[Be that flipping a heads, rolling a four, or observing rain on a given day.] actually occurs in the long run. For a Frequentist, you imagine yourself performing the experiment you are running over, and over, and over, and over, and over again. Each time you answer "did the event happen?" and you count up those occurrences. As you do this more and more and more, wherever that proportion lands corresponds to the probability.

```{r}
#| echo: false
#| label: fig-plot
#| fig-cap: "This plot simulates the repeated tossing of a coin. The x-axis represents the number of coins being tossed, and the y-axis plots the proportion of times that heads has shown up cumulatively over the tosses thus far. We can see in the long run that this proportion tends towards $0.5$."

set.seed(31415)

draws <- 5000
sequence <- rbinom(draws, 1, 0.5)

x <- 1:draws
y <- cumsum(sequence)/x

plot(y ~ x, main = "Long run proportion of heads observed tossing a fair coin.", ylab = "Proportion", xlab = "Number of Tosses", type = 'l')
abline(h=0.5, lty=3)
```

To formalize this mathematically, we first define several important terms. 

:::{#def-experiment}
## Experiment 
Any action to be performed whose outcome is not (or cannot) be known with certainty, before it is performed. 
:::
:::{#def-event}
## Event [Informally]
A specified result that may or may not occur when an experiment is performed.
:::

Suppose that an experiment is able to be performed as many times as one likes, limited only by your boredom. If you take $k_N$ to represent the number of times that the event of interest occurs when you perform the experiment $N$ times, then a Frequentist would define the probability of the event as $$\text{probability} = \lim_{N\to\infty}\frac{k_N}{N}.$$ This course does not assume that you have any familiarity with calculus, and yet, this definition relies on limits, a concept taken directly from calculus. However, we will not actually require the ability to work with limits for this course. Instead, when you see a statement of the form $$\lim_{x\to\infty} f(x),$$ simply think "what is happening to the function $f(x)$ as $x$ grows and grows (off to $\infty$)?"

In practice this means that, in order to interpret probabilities, we think about repeating an experiment many, many times over. As we do that, we observe the proportion of times that any particular outcome occurs, and take that to be the defining relation for probabilities. The reason that we say the probability of flipping a heads is $0.5$ is because if we were to sit around and flip a coin^[Our experiment.] over, and over, and over again, then in the long-run we would observe a head^[Our event.] in $0.5$ of cases.

:::{#exm-prob-interp}
## Probability Interpretation

How do we interpret the statement "the probability that Sadie would have had to pay, given a head on the first toss, is $0.25$"? Recall that in the game, they toss three coins, and Sadie pays if two of them show tails.

::::{.callout .solution collapse="true"}
## Solution
This statement means that, if Sadie were to repeatedly be in the situation where one head has shown and there are two coins left to toss, then in $0.25$ of these situations (in the limit, as this is repeated an infinite number of times) will end up showing two tails.
::::
:::

Many situations in the real world are not able to be run over and over again. Think about, for instance, the probability that a particular candidate wins in a particular election. There is uncertainty there, of course, but the election can only be run once. What then? There are several ways through these types of events. 

First, we can rely on the **power of imagination**. There is nothing stopping us from envisioning the hypothetical possibility of running the election over, and over, and over, and over again. If we step outside of reality for a moment, we can ask "if we could play the day of the election many, many, many times, what proportion of those days would end with the candidate being elected?" If we say that the candidate has a 75% chance of being elected, then we mean that in $0.75$ of those imagined worlds, the candidate wins. It is crucial to stress that in our imagination here, we need to be thinking about the **exact same day** over and over again. We cannot imagine a different path leading to the election, different speeches being given in advance, or different opposition candidates. If we start from the same place, and play it out many times over, what happens in each of those worlds?

This repeated imagining is not for everyone. As a result, alternative proposals to the interpretation of probability have been made. Most popularly, the **Bayesian interpretation** has recently become prominent. To Bayesians, probability is a measure of subjective belief. To say that there is a $50\%$ chance of a coin coming up is a statement about one's knowledge of the world. Typically, coins show up heads half the time, so that's our belief about heads. The Bayesian view, built around subjective confidence in the state of the world, can be formalized mathematically as well. A Bayesian considers the *prior evidence* that they have about the world^[Any relevant evidence that has previously been collected.] and combine this with current observations in order to update their subject beliefs, balancing these two sources of information. 

:::{.remark}
## Bayesian Probabilities and Belief Updating

Suppose that a Bayesian is flipping a coin. Before any flips have been made the Bayesian understandably believes that the coin will come up heads $50\%$ of the time. However, when the coin starts to be flipped, the observations are a string of tails in a row.

After having flipped the coin five times, the individual has observed five tails. Of course, it is totally possible to flip a fair coin five times and see five tails, but there is a level of skepticism growing.

After 10 flips, the Bayesian has still not seen a head. At this point, the subjective belief is that there is likely something unfair about this coin. Even though the experiment started with a baseline assumption that the coin was fair the Bayesian no longer believes that the next flip will be a head.

As this goes on, you can imagine the Bayesian continuing to update their view of the world. To them, the probability is an evolving concept, capturing what was believed and what has been observed.
:::

For the election example, the Bayesian interpretation is somewhat easier to think through. To say that a candidate has a $75\%$ chance of winning the election means that "based on everything that has been observed, and any prior beliefs about the viability of the candidates, the subjective likelihood that the candidate wins the election is $0.75$". If we disagree about prior beliefs, or have experienced different pieces of information, then we may disagree on the probability. That is okay.

In this course, we focus on the Frequentist interpretation. This is, in part, because Frequentist probability is an easier introduction to the concepts that are necessary for grasping uncertainty. Additionally, there is some research to suggest that Frequentist interpretations are fairly well understood by the general public ^[See, for instance @rainPaper, where they study how people interpret the statement "there is a $30\%$ chance of rain tomorrow." Interestingly, most people can convert this into a frequency statement ($3$ out of $10$, say), even if the specific meaning is sometimes lost. They conclude that there are other issues in attempting to understanding this statement, issues which we will address later on.]. However, it is important to know and recognize that there is a world beyond Frequentist Probability and Statistics, one which can be very powerful once it is unlocked. ^[More on this in later years, if you so desire! Come have a chat, if that sounds interesting.]

If probability measures the long term proportion of times that a particular event occurs, how can we go about computing probabilities? Do we require to perform an experiment over and over again? Fortunately, the answer is no. The tools of probability we will cover allow us to make concrete statements about the probabilities of events without the need for repeated experimental runs. However, before dismissing the idea of repeatedly running an experiment at face value, it is worth considering a tool we have at our disposal that renders this more possible than it has ever been: computers.

## R Programming for Probability and Statistics
Throughout this course we will make use of a programming language for statistical computing, called R. Classically, introductory statistics courses involved heavy computation of particular quantities by hand. The use of a programming language (like R) frees us from the tedium of these calculations, allowing for a deeper focus on understanding, explanation, decision-making, and complex problem solving. This is **not** to say we will *never* do problems by hand^[To the contrary, some amount of by-hand problem solving helps to solidify these concepts.], however, we will emphasize the use of statistical computing often. While you will not be expected to write R programs on your own, you will be expected to read simple scripts, make basic modifications, and to run code that is given to you.

::: {.content-visible when-format="html"}
Throughout these course notes, where relevant, R code will be provided to demonstrate the ideas being discussed. When viewing the notes on the web, you will have the capacity to actually run the code, directly in your browser, and play around with the values that it gives. The code running will be somewhat limited, and so there may be cases where you are better copying the code onto your own computer, but this provides a really excellent method for getting used to working with R code directly. In this section we will cover some of the very basics of using R, and reading R code. If you are interested, there are plenty of resources to becoming a more proficient R programmer. This is a skill that will benefit you not only in this course, but in many courses to come, and far beyond your university training. If you have any programming in your background, R is a fairly simple language to learn; if not, R can be quite beginner friendly. 
:::
::: {.content-visible when-format="pdf"}
Throughout these course notes, where relevant, R code will be provided to demonstrate the ideas being discussed. It may be useful to have R open alongside the notes to ensure that you can get the same results that are printed throughout. In this section we will cover some of the very basics of using R, and reading R code. If you are interested, there are plenty of resources to becoming a more proficient R programmer. This is a skill that will benefit you not only in this course, but in many courses to come, and far beyond your university training. If you have any programming in your background, R is a fairly simple language to learn; if not, R can be quite beginner friendly. 
:::

### Basic Introduction to R Programming
When programming, the basic idea is that we are going to write instructions in a **script** which we will tell our computer to execute. These instructions are^[Typically. There are some exceptions to this, but if this is your first time programming, you need not worry about that!] performed one-by-one, from the top to the bottom of the script. We can have instructions which operate on their own, or which interact with previous (or future) instructions to add to the complexity. The trick with programming then is to determine which actions you need the computer to perform, in which order, to accomplish the task that you are setting out to do.

To begin, we may consider a very simple R program, one which uses the programming language as a basic calculator. 

::: {.content-visible when-format="pdf"}
```{r}
5 + 3 - (10*2) + 8^(25/3)
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
5 + 3 - (10*2) + 8^(25/3)
```
:::

Here, we ask the computer to perform some simple arithmetic operations. We use `+` for addition, `-` for subtraction, `*` for multiplication, `/` for division, and `^` for exponentiation. With the use of parentheses, any expressions relating to these basic operations can be performed. Note that here the result is simply output after it is computed. Try modifying the exact expressions being computed, allowing you to feel comfortable with these types of mathematical operations.


::: {.content-visible when-format="pdf"}
```{r}
5 + 3 - (10*2) + 8^(25/3)
8 * 5
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
5 + 3 - (10*2) + 8^(25/3)
8 * 5
```
:::

Here, we have two lines of math running with simple operations. Each is simply output when it is computed. These two results have no ability to interact with one another, and if we were to add more and more lines beneath, the same would continue to happen. If we want different commands to be able to interact with one another, we need a method for storing the results. To do so, we can define **variables**. In R, to define a variable, we use the syntax `variable_name <- variable_value`. We can choose *almost* anything that we want for the variable name^[The variable name must start with a letter and can be a combination of letters, numbers, periods, and underscore.] and the variable value can also be of many types.^[more on this later] The arrow between the two is the **assignment operator** and it simply tells R to assign the `variable_value` to be accessible from the `variable_name`. 

::: {.content-visible when-format="pdf"}
```{r}
my_5 <- 5
my_8 <- 8
my_5
my_8
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_5 <- 5
my_8 <- 8
my_5
my_8
```
:::
In this code we assign the variable `my_5` to contain the value `r my_5`, and the variable `my_8` to contain the value `r my_8`. We can output these as expressions themselves, simply by typing the variable name. Simply outputting these variables is not of particular interest, however, we can use the variables in later statements by simply including the variable name in them.

::: {.content-visible when-format="pdf"}
```{r}
my_5 <- 5
my_8 <- 8
my_5*my_8
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_5 <- 5
my_8 <- 8
my_5*my_8
```
:::
Here, instead of simply outputting the variables, we multiply them together. We could have used `5 * 8` in this case for the same result, however, we are afforded a lot more flexibility with this approach. Much of this flexibility comes from our capacity to *change* the values of variables over time. Consider the following script, and try to understand why the output is the way that it is.

::: {.content-visible when-format="pdf"}
```{r}
my_5 <- 5
my_8 <- 8
my_5*my_8

my_5 <- 10
my_5*my_8
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_5 <- 5
my_8 <- 8
my_5*my_8

my_5 <- 10
my_5*my_8
```
:::
At the top point in the script, before the first `my_5*my_8` call, The variable `my_5` has the value `5`. However, after this is called, the value is updated to be `10`. Then, when we call `my_5*my_8` again, this is now simplified to `10 * 8`, giving the result. Perhaps more importantly, we can take the value of an expression and assign that to a variable itself. 

::: {.content-visible when-format="pdf"}
```{r}
my_5 <- 5
my_8 <- 8
result <- my_5*my_8
result 
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_5 <- 5
my_8 <- 8
result <- my_5*my_8
result 
```
:::
Here, we define another variable. This time, `result` now contains the result of multiplying our previous two variables. Thus, when we output it, we get the same value. Take a moment to read through the following script, and try to understand what will happen at the end. 

::: {.content-visible when-format="pdf"}
```{r}
#| output: false 
my_5 <- 5
my_8 <- 8
result <- my_5*my_8

my_5 <- result
my_8 <- my_5
result <- my_5 * my_8

result 
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_5 <- 5
my_8 <- 8
result <- my_5*my_8

my_5 <- result
my_8 <- my_5
result <- my_5 * my_8

result 
```
:::
The result is `r result`. Why? We can read through this step-by-step in order to understand this. First we set our variables to have the values of `5` and `8`. Then, `result` is made to be the product of our two variables, which in this case is `40`. After that, we set the value of `my_5` to be the same as the value of `result`, which gives `my_5` equal to `40`. At this point, `result` equals `40`, `my_5` equals `40`, and `my_8` is equal to `8`. The next line updates `my_8` to be the same as the value of `my_5`, which we just clarified was `40`. As a result, all of the variables we have defined now take on the value of `40`. The final line before output, `result <- my_5 * my_8` updates the value of `result` to be the product of the two variables again, this time giving `40 * 40` which gives `r result`. 

### Function Calls in R
Up until this point we have simply used numerical operations and variable assignment. While this allows R to serve as a very powerful calculator, we often want computers to do much more than arithmetic. As a result, we need to explore **functions** in R. A function is a piece of code which takes in various arguments and outputs some value (or values). Most of the way that we will use R in this course is through the use of function calls.^[Note: when you begin to write programs for yourself, a lot of your time will be spent writing custom functions. If this is of interest to you, I suggest looking into programming more! For this course we will not need to define our own functions, except perhaps ones that will be defined for you.] This is exactly analogous to a mathematical function: it is simply some rule which maps input to output. In fact, some of the most basic functions in R are functions which relate to mathematical functions. 

::: {.content-visible when-format="pdf"}
```{r}
x <- 10 

exp(x)  # <1>
sqrt(x) # <2>
log(x)  # <3>

```
1. Computes the exponential function applied to `x`, that is $e^x = e^{10}$.
2. Computes the square root of `x`, that is $\sqrt{x} = \sqrt{10}$.
3. Computes the natural logarithm of `x`, that is $\log(x) = \log(10)$.
:::
::: {.content-visible when-format="html"}
```{webr-r}
x <- 10 

exp(x)  # Computes the exponential, e to the power of x.
sqrt(x) # Computes the square root of x.
log(x)  # Computes the natural logarithm of x.
```
:::
The basic format of a function call will always be `function_name(param1, param2, ...)`. Each of these functions required only a single parameter, however, there are some functions which take more than one parameter. If we have decimal numbers, for instance, we may wish to round them. To do so, we can use the `round` function in R, which takes two parameters: the number we wish to round, and how many digits we wish to keep. 

::: {.content-visible when-format="pdf"}
```{r}
unrounded_value <- 3.141592
rounded_value <- round(unrounded_value, digits = 3)

rounded_value
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
unrounded_value <- 3.141592
rounded_value <- round(unrounded_value, digits = 3)

rounded_value
```
:::
There are a few things to note about this sequence of function calls. First, note that we assign the output of a function to a new variable. This behaves exactly as we saw above with simple numeric calculations. Next, consider that the output of the function^[Which we have stored in the variable `rounded_value`] has a value of $3.142$. That is: we rounded the value to $3$ decimal points, exactly as would be expected. The final part to note is that the second parameter passed to the function call is **named**. That is, instead of simply calling `round(unrounded_value, 3)`, we have `round(unrounded_value, digits = 3)`. If you had made the first call this would have worked perfectly.^[Try it out to convince yourself!] However, R also provides the ability to pass in parameter names alongside the parameter values with the syntax `function_name(param_name = param_value, ...)`. The benefit to doing this is two-fold. First, it is easier to read what is happening, especially for function calls that you have never seen before. Second, it removes the need to have the parameters ordered correctly. It is best practice to **always** include parameter names where you can. 

Now, you may be wondering "how do you know what the parameter names should be?" The names are built-in to the different functions that you are working with and so overtime you will become quite familiar with them. However, at any point you can also run the command `?function_name`, replacing `function_name` with the name of a function you are interested in, to open documentation about that function. There you will see not only the names of the different parameters, but useful information regarding what the function does, examples of how to call it, and so forth. 

::: {.content-visible when-format="pdf"}
```{r}
#| eval: false
?round
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
?round
```
:::
When we run this code, we are given *a lot* of information. We can see the function name, the details, how it's called, and so forth. In the **Arguments** section we get a list of all of the arguments we can pass to the function, along with a description of them. In this case we see that `round` can take a parameter called `x` for the number to be rounded, and `digits` (which we have previously seen).

::: {.content-visible when-format="pdf"}
```{r}
unrounded_value <- 3.141592
rounded_value <- round(digits = 3, x = unrounded_value)

rounded_value
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
unrounded_value <- 3.141592
rounded_value <- round(digits = 3, x = unrounded_value)

rounded_value
```
:::
This code produces the exact same output, where now both parameters are named. Specifically, we could read this function call as "round the value of `x` to have `digits` decimal points." If you had instead written `round(3, unrounded_value)`, you would get the value `3`, since now we are rounding `3` to `3.141592` decimal points. 

### Moving Beyond Numeric Data
So far, everything that we have looked at has been numeric data. We have seen integers, and decimals. You can have negative results, say by taking `my_var <- -5`. And while numbers are frequently useful, we will require further types of data to write useful computer programs. For this course we will focus on three additional data types: textual data which (referred to as **strings**), true and false binary data (referred to as **booleans**), and lists of the same data type (referred to as **vectors**). They will behave in much the same way as numeric data, with different functions and techniques which can be applied to them. 

To define a string of text, we simply encapsulate the text that we are interested in within quotation marks (either single `'` or double `"` quotation marks will work). 

::: {.content-visible when-format="pdf"}
```{r}
first_string <- "This is a string."
second_string <- 'This is also a string.'

first_string
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
first_string <- "This is a string."
second_string <- 'This is also a string.'

first_string
```
:::
Two commonly used functions which rely on strings are `paste` and `print`. Each will take in strings as input, and they do not need to be named. The function `paste` can take in as many strings as you would like. It will "paste" together all the strings provided, creating a longer string out of these. The function `print` will display the string that is passed as output. Until now we have been running these programs in a way where all calls are displayed as output: this will not always be the case, and so print can come in handy there. 

::: {.content-visible when-format="pdf"}
```{r}
my_greeting <- "Hello! Welcome to R programming,"
my_name <- "Dylan"

combined_string <- paste(my_greeting, my_name, "!")
print(combined_string)
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_greeting <- "Hello! Welcome to R programming,"
my_name <- "Dylan"

combined_string <- paste(my_greeting, my_name, "!")
print(combined_string)
```
:::
Note that `paste` has several additional options which can be investigated in the documentation for `paste`. This is only the simplest use case for it. In general, strings are particularly helpful when we wish to have output from the computer that will be human readable. Where strings are largely for humans, booleans are largely for computers.

Much of what computer programming entails is checking whether certain conditions hold, and then taking different actions depending on what is found. In order to do this, the computer needs a way to represent true and false statements. In R, these are codified with the values `TRUE` and `FALSE`. Note, the capital letters here make a difference. You cannot use `true` or `True` or any other combination thereof. More important than simply being able to specify the values `TRUE` and `FALSE` directly is the ability to detect whether certain statements are `TRUE` or `FALSE`. For this we require comparison operators.

If we think of mathematical comparisons we can state whether two things are equal, or not equal, and whether one thing is less than (or equal to) or greater than (or equal to) another. We can run all of these same checks in R. 

* To check whether two quantities are equal you use `quantity_1 == quantity_2`. This statement will be `TRUE` if `quantity_1` and `quantity_2` are exactly the same, and will be `FALSE` otherwise. 
* To check whether two quantities are not equal, you can use `quantity_1 != quantity_2`. This statement will be `TRUE` if the quantities differ from one another. 
* To check whether one quantity is larger than another, you can use `quantity_1 > quantity_2`. If you want to know whether it is greater than *or* equal to, you can use `quantity_1 >= quantity_2`. 
* To check whether one quantity is smaller than another, you can use `quantity_1 < quantity_2`. If you want to know whether it is less than *or* equal to, you can use `quantity_1 <= quantity_2`. 

::: {.content-visible when-format="pdf"}
```{r}
5 == 5 # TRUE
5 == 6 # FALSE

5 != 5 # FALSE 
5 != 6 # TRUE 

5 > 6 # FALSE
5 > 5 # FALSE

5 >= 6 # FALSE
5 >= 5 # TRUE

5 < 6 # TRUE
5 < 5 # FALSE 

5 <= 6 # TRUE 
5 <= 5 # TRUE 
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
5 == 5 # TRUE
5 == 6 # FALSE

5 != 5 # FALSE 
5 != 6 # TRUE 

5 > 6 # FALSE
5 > 5 # FALSE

5 >= 6 # FALSE
5 >= 5 # TRUE

5 < 6 # TRUE
5 < 5 # FALSE 

5 <= 6 # TRUE 
5 <= 5 # TRUE 
```
:::
There are two key points to note beyond this. First, we will of course not normally compare two constants to one another. We already know that `5==5` and so we would not need to check it. We can, however, plug-in variables, perhaps with unknown values, and have the same types of statements being made. Second, the checks for equality and inequality also work with other data types (like strings).

::: {.content-visible when-format="pdf"}
```{r}
string1 <- "STRING1"
string2 <- "STRING1"
string3 <- "string1"
string4 <- "5"
string5 <- "5.3421"
num1 <- 5
num2 <- 1
num3 <- 0
bool1 <- TRUE 
bool2 <- FALSE 

string1 == string2 # TRUE
string1 == string3 # FALSE 
string2 != string3 # TRUE
string1 != string4 # TRUE 

string4 == num1    # TRUE 
string1 == num2    # FALSE
num1 == string5    # FALSE
num2 == bool1      # TRUE 
bool2 == num3      # TRUE 
bool2 == num2      # FALSE
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
string1 <- "STRING1"
string2 <- "STRING1"
string3 <- "string1"
string4 <- "5"
string5 <- "5.3421"
num1 <- 5
num2 <- 1
num3 <- 0
bool1 <- TRUE 
bool2 <- FALSE 

string1 == string2 # TRUE
string1 == string3 # FALSE 
string2 != string3 # TRUE
string1 != string4 # TRUE 

string4 == num1    # TRUE 
string1 == num2    # FALSE
num1 == string5    # FALSE
num2 == bool1      # TRUE 
bool2 == num3      # TRUE 
bool2 == num2      # FALSE
```
:::
The final checks may be slightly odd. Here we are comparing across different types of data. When we do this R will automatically try to convert from one type to the other. With strings and numbers this is not too challenging. If they can be converted nicely between types, then they are and the values are compared. Otherwise, R will conclude they are not equal by default. For booleans, it is important to note that `TRUE == 1` and `FALSE == 0`. We will often use these values interchangeably. 

The final data type that we will consider are vectors. Vectors store many different values, of the same type, in a single object. Thus, we may have a vector of numeric data, or a vector of strings, or a vector of booleans. The vectors will always contain the same type throughout, but they are stored in a single object (and as such, for instance, can be stored in a single variable). To define a vector we call `c(...)`, where the `...` contains the set of objects we want to store in the vector. The `c` stands for **c**oncatenate, as we are *concatenating* together the set of items into a single container. 

::: {.content-visible when-format="pdf"}
```{r}
v1 <- c("vector", "of", "strings")
v2 <- c(3, 1, 4, 1, 5)
v3 <- c(TRUE, TRUE, FALSE, TRUE)
v4 <- c(1==2, 2==2, 3==4)
v1
v2
v3
v4
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
v1 <- c("vector", "of", "strings")
v2 <- c(3, 1, 4, 1, 5)
v3 <- c(TRUE, TRUE, FALSE, TRUE)
v4 <- c(1==2, 2==2, 3==4)
v1
v2
v3
v4
```
:::
We see that each of these vectors holds one type of object. Vectors can be of arbitrary and different lengths. It is also possible to combine multiple vectors *of the same type* into one, by using the `c` function again. 

::: {.content-visible when-format="pdf"}
```{r}
v1 <- c(3,1,4,1,5)
v2 <- c(9,2,6)
v3 <- c(5)
num1 <- 3

combined_v1 <- c(v1, v2)
combined_v2 <- c(combined_v1, v3)
combined_v3 <- c(combined_v2, num1)

combined_v3
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
v1 <- c(3,1,4,1,5)
v2 <- c(9,2,6)
v3 <- c(5)
num1 <- 3

combined_v1 <- c(v1, v2)
combined_v2 <- c(combined_v1, v3)
combined_v3 <- c(combined_v2, num1)

combined_v3
```
:::
Here we combine different numeric vectors together. We also show, when forming `combined_v3`, how numeric vectors can have single items added onto them. That is, if you have a single number, it can be treated as a vector with one element in it. This becomes very useful when building up vectors within code. In addition to combining multiple vectors together, we can also select elements out of a vector. To do this, we use a set of square brackets after the vector's name, with a number within those square brackets specifying the **index** of the vector we are interested in. The index is simply the element position starting at 1^[If you have programmed in the past there is a good chance the language you have learned is "0 indexed" rather than "1 indexed". In R, all vectors start at position 1 and count up, which is not the case in many languages. Be careful of this.] and running to the length of the vector. We can include a vector of indices to select multiple elements at once.

::: {.content-visible when-format="pdf"}
```{r}
alpha_vector <- c("A","B","C","D","E","F","G","H","I",
                  "J","K","L","M","N","O","P","Q","R",
                  "S","T","U","V","W","X","Y","Z")

alpha_vector[4]    # D
alpha_vector[25]   # Y 
alpha_vector[12]   # L
alpha_vector[1]    # A
alpha_vector[14]   # N

my_name <- alpha_vector[c(4,25,12,1,14)]

my_name 
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
alpha_vector <- c("A","B","C","D","E","F","G","H","I",
                  "J","K","L","M","N","O","P","Q","R",
                  "S","T","U","V","W","X","Y","Z")

alpha_vector[4]    # D
alpha_vector[25]   # Y 
alpha_vector[12]   # L
alpha_vector[1]    # A
alpha_vector[14]   # N

my_name <- alpha_vector[c(4,25,12,1,14)]

my_name 
```
:::
Note that, each element is selectable individually, giving a single item of that type (in this case, strings). If you select multiple of the elements together, it will create a vector of that type (in this case, a string vector). Note that in addition to selecting elements in this way by their indices, you can also update the elements in the same way. 

::: {.content-visible when-format="pdf"}
```{r}
cur_year <- c(2,0,2,3)
cur_year 

# After Midnight on December 31
cur_year[4] <- 4
cur_year 
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
cur_year <- c(2,0,2,3)
cur_year 

# After Midnight on December 31
cur_year[4] <- 4
cur_year 
```
:::
In this example we are changing the last element of the vector. Sometimes we may not know how long the vector actually is, if for instance, it is being built-up as our code runs. If we ever want to check the length of a vector, we can simply call the function `length` which takes as input only one vector, and outputs the numeric value of its length. 

::: {.content-visible when-format="pdf"}
```{r}
v1 <- c(1,2,3,4,5)
length(v1)
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
v1 <- c(1,2,3,4,5)
length(v1)
```
:::

### Program Control Flow
We have seen different types of data, different ways of manipulating data, functions, and variables so far. In order to bring all of these concepts together into useful programs we need some way to control the flow of our programs. We have seen that, by default, programs execute from the top until the bottom. However, it will often be the case that we want want to have certain code running only if certain conditions hold, or that we want to repeat some piece of code many times over. To accomplish these tasks we require **control flow statements**. We will consider only two types of control flow statements now, which will serve well enough to read most of what needs to be read for this course. 

The first type of statement is the **conditional statement**. Conditional statements execute only when certain conditions hold. The simplest conditional statement is an **if** statement. The format to define an if statement if `if (condition){ ... }`, where `condition` is some logical condition to be evaluated. If the condition is `TRUE` then the code contained in { ... } is evaluated. If the condition is `FALSE` it is not. 

::: {.content-visible when-format="pdf"}
```{r}
my_number <- 5

if(my_number > 0) {
    print("My number is a positive.")
}

if(my_number < 0) {
    print("My number is a negative.")
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_number <- 5

if(my_number > 0) {
    print("My number is a positive.")
}

if(my_number < 0) {
    print("My number is a negative.")
}
```
:::
In this case, these simple conditional statement check to see whether the number we entered is larger than zero and whether the number we entered is smaller than zero, respectively. When run, notice that only one of the statements is executed.^[In fact, if `my_number` is set to `0` then none of the statements are executed.] In this case, we know that only one (or neither) of these statements can be true. When that is the case it may make sense to make use of `else` clauses in our conditional logic.

::: {.content-visible when-format="pdf"}
```{r}
my_number <- -5

if(my_number > 0) {
    print("My number is a positive.")
} else {
    print("My number is not a positive.")
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_number <- -5

if(my_number > 0) {
    print("My number is a positive.")
} else {
    print("My number is not a positive.")
}
```
:::
Here, *if* the number is greater than `0`, then we run the first block of code, *otherwise* we run the second block of code. Thus, whenever a positive number is entered, we see "My number is a positive", and whenever a non-positive number is entered, we see "My number is not a positive." We can extend `else` blocks to be `else if` blocks, where further conditions can be specified. 

::: {.content-visible when-format="pdf"}
```{r}
my_number <- 100

if(my_number > 50) {
    print("My number is a large positive value.")
} else if(my_number > 0){
    print("My number is a small positive value.")
} else if(my_number < -50) {
    print("My number is a large negative value.")
} else if(my_number < 0) {
    print("My number is a small negative value.")
} else {
    print("My number is zero.")
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_number <- 100

if(my_number > 50) {
    print("My number is a large positive value.")
} else if(my_number > 0){
    print("My number is a small positive value.")
} else if(my_number < -50) {
    print("My number is a large negative value.")
} else if(my_number < 0) {
    print("My number is a small negative value.")
} else {
    print("My number is zero.")
}
```
:::
In these case we can simply pass through each conditional statement in order. First, is the number larger than `50`? If so, print the statement, otherwise we check the next condition, is the number greater than `0`? Note that if we are checking this condition we *know* that the number is less than or equal to `50` since it failed the first check. We continue through the rest of this procedure down until the last else block. This block is run only when all of the other conditions fail: that is, our value is not larger than `50`, or larger than `0`, or smaller than `-50`, or smaller than `0`. The only value that satisfies this is `0` itself. 

Sometimes, we wish to check compound conditions. That is, we want to know whether multiple conditions hold, or perhaps, whether at least one of many conditions hold. These statements can be converted into "and" and "or" statements, respectively. To denote "and" statements we use `&&` and to denote "or" statements we use `||`. Thus, the check `my_val > 0 && my_val < 100` returns true only if `my_val` is both above `0` **and** below 100. The check `my_val < -50 || my_val > 50` returns true whenever **either** `my_val` is less than `-50` **or** `my_val` is greater than `50`. 

::: {.content-visible when-format="pdf"}
```{r}
my_number <- 5

if(my_number < 50 && my_number > 10) {
    print("A moderate, positive number.")
} else if(my_number != 5 && my_number <= 10 && my_number >= 0) {
    print("A positive value which is not 5.")
} else if(my_number == 5 || my_number < 0) {
    print("Either 5 or a negative value.")
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
my_number <- 5

if(my_number < 50 && my_number > 10) {
    print("A moderate, positive number.")
} else if(my_number != 5 && my_number <= 10 && my_number >= 0) {
    print("A positive value which is not 5.")
} else if(my_number == 5 || my_number < 0) {
    print("Either 5 or a negative value.")
}
```
:::
Conditional statements can grow to be very complex, however, with these rules you can read through them top-to-bottom, substituting for "and" and "or" where necessary. It is also possible, where required, to place one conditional statement inside the code block for another, and to combine them with any of the other techniques that we have learned thus far. 

The final piece of control flow that we will consider for now is the `for` loop. The idea with a `for` loop is that we want to repeat the same action either a certain number of times, or for every item in a set of items. To do so, we use the syntax `for(x in vector){...}`, where the code in `...` will be performed once for every single item in the vector. Within the code block specified by `...`, the value `x` will take on the current value in the loop. 

::: {.content-visible when-format="pdf"}
```{r}
for(x in c(1,2,3)){
    print(x)
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
for(x in c(1,2,3)){
    print(x)
}
```
:::
Notice that three values are printed, in order, `1` then `2` then `3`. The loop code is run three different times, one for each element in the list. Each time the loop is running the next value from the list gets assigned to the variable `x`. The first time it runs it gets the first element, and so forth. As a result, we can use these values in our calculations in whatever way we need to. 

::: {.content-visible when-format="pdf"}
```{r}
for(x in c(1,2,3)){
    x_sq <- x^2
    print(paste("The square of", x, 'is', x_sq))
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
for(x in c(1,2,3)){
    x_sq <- x^2
    print(paste("The square of", x, 'is', x_sq))
}
```
:::
Whenever we are trying to form a numeric vector with consecutive elements, as we are in `c(1,2,3)`, we can make this easier on ourselves by simply specifying the upper and lower bounds of the range, separated by a colon. That is `c(1,2,3) == 1:3`. This is often useful when specify a loop as we very often want to repeat something a set number of times. Note that we do not ever *need* to use the value of the looping variable. Sometimes, we just want things to repeat, and so the loop is a convenient way to do that. 

::: {.content-visible when-format="pdf"}
```{r}
times_to_loop <- 5

for(x in 1:times_to_loop){
    print(paste("This will get printed", times_to_loop, "times."))
}
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
times_to_loop <- 5

for(x in 1:times_to_loop){
    print(paste("This will get printed", times_to_loop, "times."))
}
```
:::

### Reading Through a More Complex R Program
Take a moment to read through the following R program and try to understand what is happening exactly. There are comments throughout which will assist in the parsing of the script. We have seen comments up until now, without drawing explicit attention to them. In R, anything placed after a `#` on the line is considered a comment. The programming language ignores these and so they are only there to help other individuals who may be reading through. It is good practice to comment your code to help others, and also to help yourself whenever you return to it in the future. In the course notes code will typically be commented. If you are reading the PDF version of the notes often these comments will be annotations beside the code (numbering certain lines) with the comments provided below, for the sake of legibility.

Note, this combines everything that we have learned, and it is entirely understandable if it takes some time to process. Fortunately, you can always try running the script yourself, and playing with different components of it. Remember, if you do not know what a function call does, you can use the documentation^[The internet is also a wonderful resource, one which even very experienced developers make frequent use of.] and try playing with it some yourself. To help with the interpretation here, note that this is an implementation of the game that Charles and Sadie have been playing.

::: {.content-visible when-format="pdf"}
```{r}
# Define some variables which dictate how the game runs.
player1 <- "Charles"
player2 <- "Sadie"
num_of_flips <- 3
flip_option1 <- "H"
flip_option2 <- "T"

# Begin with the game setup
num_to_win <- round(x = num_of_flips / 2, 0)
flip_results <- c()
flip_options <- c(flip_option1, flip_option2)
total_1 <- 0
total_2 <- 0
player_name <- "" # This is an 'empty string'

# Start Playing the Game
for(game_round in 1:num_of_flips) {
    print(paste("Now starting round", game_round))

    # Select a flip of the coin, using the sample function.
    # See more details with ?sample
    flip_result <- sample(x = flip_options, size = 1)

    # See who benefits from this flip
    # Then set player_name to be the player who benefits from the flip,
    # and update their score variable.
    if (flip_result == flip_option1) {
        player_name <- player1
        total_1 <- total_1 + 1
    } else if (flip_result == flip_option2) {
        player_name <- player2
        total_2 <- total_2 + 1
    }

    print(paste("The flip was a", flip_result, "which benefits", player_name))

    # Check to see if either player has won at this point.
    if(total_1 >= num_to_win) {
        print(paste(player1, "has scored enough points to win."))
    } else if(total_2 >= num_to_win){
        print(paste(player2, "has scored enough points to win."))
    }
}

# The game is over
# Select the winner based on who scored above the threshold 
# and then print out the results.
winner <- player1 

if(total_2 >= num_to_win) {
    winner <- player2 
}

# Separate these statements into multiple lines to ensure the text is 
# not too wide.
print(paste("After flipping the coin", num_of_flips, "times,", player1, 
            "scored a total of", total_1, "points"))
print(paste("while", player2, "scored a total of", total_2, "points."))
print(paste("As a result", winner, "won the game and will not have to pay!"))
```
:::
::: {.content-visible when-format="html"}
```{webr-r}
# Define some variables which dictate how the game runs.
player1 <- "Charles"
player2 <- "Sadie"
num_of_flips <- 3
flip_option1 <- "H"
flip_option2 <- "T"

# Begin with the game setup
num_to_win <- round(x = num_of_flips / 2, 0)
flip_results <- c()
flip_options <- c(flip_option1, flip_option2)
total_1 <- 0
total_2 <- 0
player_name <- "" # This is an 'empty string'

# Start Playing the Game
for(game_round in 1:num_of_flips) {
    print(paste("Now starting round", game_round))

    # Select a flip of the coin, using the sample function.
    # See more details with ?sample
    flip_result <- sample(x = flip_options, size = 1)

    # See who benefits from this flip
    # Then set player_name to be the player who benefits from the flip,
    # and update their score variable.
    if (flip_result == flip_option1) {
        player_name <- player1
        total_1 <- total_1 + 1
    } else if (flip_result == flip_option2) {
        player_name <- player2
        total_2 <- total_2 + 1
    }

    print(paste("The flip was a", flip_result, "which benefits", player_name))

    # Check to see if either player has won at this point.
    if(total_1 >= num_to_win) {
        print(paste(player1, "has scored enough points to win."))
    } else if(total_2 >= num_to_win){
        print(paste(player2, "has scored enough points to win."))
    }
}

# The game is over
# Select the winner based on who scored above the threshold 
# and then print out the results.
winner <- player1 

if(total_2 >= num_to_win) {
    winner <- player2 
}

print(paste("After flipping the coin", num_of_flips, "times,", player1, 
            "scored a total of", total_1, "points, while", player2, 
            "scored a total of", total_2, "points. As a result",
            winner, "won the game and will not have to pay!"))
```
:::


### R Programming for Probability Interpretations
Recall that the motivation for the discussion of R was the Frequentist interpretation of probability. One task that computers are very effective at is repeatedly performing some action. As a result, we can use computers to mimic the idea of repeatedly performing an experiment. Consider the simple case of flipping a coin over and over again.

We can use `sample(x, size)` as a function to select `size` realizations from the set contained in `x`. Thus, if we take `sample(x = c("H","T"), size=1)` we can view this as flipping a coin one time. If we use the loop structure we talked before, then we can simulate the experience of repeatedly flipping a coin. Consider the following R code.

::: {.content-visible when-format="pdf"}
```{r}
set.seed(3141592) # <1> 

number_of_runs <- 1000 # <2> 
tosses <- c() # <3> 

for(idx in 1:number_of_runs){ # <4> 
  toss <- sample(x = c("H","T"), size = 1) # <5> 
  if(toss == "H"){ # <6> 
    tosses <- c(tosses, 1) # <6> 
  } else { # <6> 
    tosses <- c(tosses, 0) # <6> 
  }
}

mean(tosses) # <7> 
```
1. A seed ensures that the random numbers generated by the program are always the same. This helps to be able to reproduce our work.
2. This is how many times we want to repeat the experiment.
3. This is where we are going to store the results of our tosses. It creates an empty list for us.
4. Here we are going to loop over the experiments, one for each run.
5. This is our coin toss. We are going to sample 1 from either 'H' or 'T'
6. If the coin toss is heads, then we add a 1 to the list. Otherwise, we add a zero to the list.
7. Return the mean of all of the tosses.
:::
::: {.content-visible when-format="html"}
Consider the following code block and this time actually see if you can update the sample sizes, or change the generation process at all. There is no harm in trying: anything you do can be undone simply be reloading the page! 
```{webr-r}
# A seed ensures that the random numbers generated by the program 
# are always the same. This helps to be able to reproduce our work.
set.seed(3141592) 

# This is how many times we want to repeat the experiment.
number_of_runs <- 1000 

# This is where we are going to store the results of our tosses. 
# It creates an empty list for us.
tosses <- c() 

# Here we are going to loop over the experiments, one for each run.
for(idx in 1:number_of_runs){ 
    # This is our coin toss. We are going to sample 1 from either 
    # 'H' or 'T'
    toss <- sample(x = c("H","T"), size = 1) 

    # If the coin toss is heads, then we add a 1 to the list. 
    # Otherwise, we add a zero to the list.
    if(toss == "H"){ 
        tosses <- c(tosses, 1) 
    } else { 
        tosses <- c(tosses, 0) 
    }
}

# Return the mean of all of the tosses.
mean(tosses) 
```
:::

It is worth adjusting some of the parameters within the simulation, and seeing what happens. What if you ran the experiment only 5 times? Ten thousand times? What if instead of counting the number of heads, we wanted to count the number of tails? What if we wanted to count the number of times that a six-sided die rolled a 4? All of these settings can be investigated with simple modifications to the provided script.

### References  {.unnumbered}