# The Mathematical Foundations of Statistical Experiments
## The Sample Space and Events
In Chapter 1 we saw the mathematical formulation for the Frequentist interpretation of probability. To study probability, we require a more detailed mathematical model. We want a description, framed in terms of mathematical objects,  which will allow us to work out probabilities of interest. In general, to form such a probability model we need both a list of all possible outcomes that the experiment can produce, as well as the probabilities of these outcomes.

We call the list of outcomes that can occur from an experiment the **sample space** of the experiment. The sample space is denoted $\mathcal{S}$, and is defined as the set of all possible outcomes from the experiment. For instance, if the experiment is flipping a coin we have $\mathcal{S} = \{\text{H}, \text{T}\}$. If the experiment is rolling a six-sided die then $\mathcal{S} = \{1,2,3,4,5,6\}$. 

:::{#def-sample-space}
## Sample Space
The sample space of a statistical experiment is the set of all possible outcomes that can be realized from that experiment. The sample space is typically denoted $\mathcal{S}$, or with similar script letters.
:::

:::{#exm-sample-space-basic}
## Enumerating Sample Spaces

Write down the complete sample space, $\mathcal{S}$ for the game that Sadie and Charles play, based on flipping and observing a coin three times in sequence.

::::{.callout .solution collapse="true"}
## Solution

For Sadie and Charles their experiment involves tossing a coin three times in sequence. As a result each outcome is a three-dimensional list of values, given for instance by $(\text{H},\text{H},\text{H})$. We can write the full sample space as $$\mathcal{S} = \{(\text{H},\text{H},\text{H}), (\text{H},\text{H},\text{T}), (\text{H},\text{T},\text{H}), (\text{H},\text{T},\text{T}), (\text{T},\text{H},\text{H}), (\text{T},\text{H},\text{T}), (\text{T},\text{T},\text{H}), (\text{T},\text{T},\text{T})\}.$$
::::
:::

With the sample space formally defined, we can revisit @def-event-inf, and formally define the concept of an event. 

:::{#def-event}
## Event 

An event is any collection of outcomes from a sample space for a statistical experiment. Mathematically, an event, $E$, is a subset of $\mathcal{S}$, and we write $E\subset\mathcal{S}$.
:::

Take for instance the experiment of a single coin. In this case, we may define $E_1 = \{\text{H}\}$, $E_2 = \{\text{T}\}$, and $E_3 = \{\text{H},\text{T}\}$ as examples of possible events. Here, $E_1$ corresponds to the event that a head is observed, $E_2$ corresponds to the event that a tail is observed, and $E_3$ corresponds to the event that either a tails or a heads was observed. Note that for each event we have $E_1 \subseteq\mathcal{S}$, $E_2 \subseteq\mathcal{S}$, and $E_3 \subseteq \mathcal{S}$.^[Note that the symbol $\subseteq$ is the *subset or equal* symbol. If we write $A \subseteq B$, then $A$ is either a subset of $B$ or else $A$ is equal to $B$. This is in contrast to using $A \subseteq B$ which suggests that $A$ is not equal to $B$. The distinction is analogous to the difference between writing $x < y$ versus $x \leq y$. Throughout these notes we will predominantly use $\subseteq$.]

:::{#exm-basic-events}
## Basic Event Listing
List several events from the game that Charles and Sadie are playing. Indicate why these are events.

::::{.callout .solution collapse='true'}
## Solution
Recall that an event is any subset of the sample space. In @exm-sample-space-basic we define $\mathcal{S}$ for this game. As a result we can take sets which contain any combinations of these elements. For instance $E_1 = \{(\text{H},\text{H},\text{H})\}$, or $E_2 = \{(\text{H},\text{H},\text{T}), (\text{H},\text{T},\text{H})\}$, or $$E_3 = \{(\text{H},\text{H},\text{H}), (\text{H},\text{H},\text{T}), (\text{H},\text{T},\text{H}), (\text{H},\text{T},\text{T}), (\text{T},\text{H},\text{H}), (\text{T},\text{H},\text{T}), (\text{T},\text{T},\text{H}), (\text{T},\text{T},\text{T})\}.$$ These are all events since $E_1 \subseteq\mathcal{S}$, $E_2 \subseteq\mathcal{S}$, and $E_3 \subseteq\mathcal{S}$.
::::
:::

:::{#exm-event-identification}
## Event Identification
Is "Charles has to pay" an event from the game that Charles and Sadie are playing? Why? Explain.

::::{.callout .solution collapse='true'}
## Solution
"Charles has to pay" is not directly an event, as it is not a subset of the sample space. This could plausibly be seen as a real-world description of a possible event, but it is not *itself* an event. The strictness of this language will be relaxed as time goes on, however, it is important to familiarize yourself with how descriptions are converted to events. In this case, we could take an event to be 
$$E = \{(\text{H},\text{H},\text{H}), (\text{H},\text{H},\text{T}), (\text{H},\text{T},\text{H}), (\text{T},\text{H},\text{H})\},$$
and state that if $E$ occurs then Charles has to pay.^[When speaking to a statistician, they would understand "Charles has to pay" as an event that *can* occur based on the defined sample space, by simply transforming it into the language of the sample space. However, the distinction is important to make: events are always subsets of the sample space. Once this is second nature, it is a rule that can be loosened, as the knowledge can always be fallen back on when needed. Simply put: you need to know the rules in order to break them!]
::::
:::

:::{#exm-event-conversion}
## Defining Events from Real-World Descriptions
What event corresponds to the description "Sadie has to pay" in the game that Charles and Sadie are playing? Recall that they flip a coin three times, and Charles will pay if at least two heads come up, while Sadie will pay if at least two tails come up.

::::{.callout .solution collapse='true'}
## Solution
Sadie will have to pay whenever there are two or more tails. As a result we can enumerate the possible outcomes that leads to Sadie paying. We have

1. $(\text{T},\text{T},\text{T})$; 
2. $(\text{H},\text{T},\text{T})$; 
3. $(\text{T},\text{H},\text{T})$; 
4. $(\text{T},\text{T},\text{H})$.

Any other outcome will have fewer than two tails, and as a result, Sadie will not have to pay. Thus, to form an event, we consider the set with each of these outcomes in it. This gives $$E = \{(\text{T},\text{T},\text{T}), (\text{H},\text{T},\text{T}), (\text{T},\text{H},\text{T}), (\text{T},\text{T},\text{H})\}.$$
::::
:::

While the events $E_1 = \{\text{H}\}$ and $E_2 = \{\text{T}\}$ each correspond to a simple outcome from the sample space, $E_3 = \{\text{H},\text{T}\}$ corresponds to a combined event. We call direct outcomes **simple events** and more complex outcomes like $E_3$ **compound events**. 

:::{#def-simple-event}
## Simple Event
A simple event is any event which corresponds to exactly one outcome from the sample space. A simple event only has one way of occurring. The size of the set for a simple event will be $1$. The sample space, in turn, is made up of a collection of simple events.
:::

:::{#def-compound-event}
## Compound Event
A compound event is any event which corresponds to more than one outcome from the sample space. A compound event can occur in multiple different ways. The size of the set for the compound event will be greater than $1$.
:::

If we consider rolling a six-sided die, then an example of a simple event is that a four shows up, denoted $\{4\}$. A compound event could be that an even number is rolled, $\{2,4,6\}$, or that a number greater  than or equal to four is rolled, $\{4, 5, 6\}$. 

:::{#exm-simple-versus-compound}
## Identifying Simple and Compound Events
List an example of (at least) one simple and one compound event from the game that Charles and Sadie are playing. 

::::{.callout .solution collapse='true'}
## Solution
An example of a simple event would be $E_1 = \{(\text{H},\text{H},\text{H})\}$ since it is comprised of exactly one outcome. If three heads are rolled, this event occurs, there is no other way for it to occur. An example of a compound event would be $$E_2 = \{(\text{H},\text{H},\text{H}), (\text{T},\text{H},\text{H}), (\text{H},\text{T},\text{H}), (\text{H},\text{H},\text{T})\}.$$ Here there are four outcomes that correspond to this event, and if any of those outcomes are observed the event occurs.
::::
:::

We say that an event "occurs" if any of the outcomes comprising the event occur. As a result we can have more than one event occurring as the result of a run of a statistical experiment. Suppose that we are rolling a fair, six-sided die. Consider the events "an even number was rolled" and "a number greater than or equal to four was rolled." If a four or a six are rolled, both of these events happen simultaneously. Our goal when working with probability will be to assign probability values to different events. We will talk about how likely, or unlikely, events of interest are, given the underlying statistical experiment.

Above, we defined $E_3$ to be equal to $\mathcal{S}$. As a result, we can say that $\mathcal{S}$ *is* an event since $\mathcal{S} \subseteq \mathcal{S}$. This is the event that any outcome is observed, which is certain to happen. Since it is certain to happen, we know it happens with probability $1$. There is another "special" event which is important to consider. We call this the *null event*. Denoted $\emptyset$, the null event is an event that corresponds to "nothing in the sample space". We know that every time an experiment is run something in the sample space occurs, and so the null event is assigned probability zero.

:::{#def-null-event}
## Null Event
The null event, denoted $\emptyset$ or $\{\}$, is an event from a statistical experiment which corresponds to nothing within the sample space. The null event has probability zero, and it is impossible to observe. Note that, no matter the sample space, $\emptyset\subset\mathcal{S}$.
:::

## Set Operations for Event Manipulation
Ultimately, all events are sets. These sets are subsets of the sample space, and can contain single or multiple outcomes. Every quantity that we are interested in can be expressed as some set of outcomes of interest. In building up these sets it is common to construct them through the use of "and", "or", and "not" statements. That is, we may say that our event occurs if some outcome **OR** another outcome occurs, or perhaps our outcomes occurs if some outcome does **NOT** occur.^[While both *or* and *not* language is likely clear from the examples we have seen so far, *and* language may be slightly less obvious. While we will explore this in more depth shortly, note that you could not have two simple events occurring simultaneously. If $E_1$ and $E_2$ are both simple events, then you can have $E_1$ **or** $E_2$, and you can have **not** $E_1$, but you cannot have $E_1$ **and** $E_2$. This is *not* true for compound events.]

Consider the example of drawing cards from a standard 52-card deck. In such a deck there are 13 card ranks, and four card suits, with one of each combination present. If we draw a single card we can think of the outcomes of the experiment as being any of the 52 possible combinations of rank and suit. We are often interested in an event such as "the card is red", which is the same as saying "the card is a heart **or** the card is a diamond." Perhaps we want to know whether the card was an ace through ten, this is the same as saying "the card is **not** a Jack **or** a Queen **or** a King." If we are interested in the event that the ace of spades was drawn, this can be expressed by saying that "the card was a spade **and** the card was an ace." 

As you begin to pay attention to the linguistic representation of events that we use, you will notice more and more the use of these words to form compound events in particular. As a result, we give each of them a mathematical operation which allow us to quickly and compactly express these quantities in notation.

:::{#exm-linguistic-description}
## Description of Events
Describe "Charles has to pay", based on the game Charles and Sadie are playing, using language revolving around "or", "and", and "not" each. That is, describe observing at least two heads, on three flips of a coin, one time using "or", one time using "and", and one time using "not".^[It may be helpful to notice that you can mix and match these terms to your hearts content!]

::::{.callout .solution collapse='true'}
## Solution
There are many possible ways of doing so. Consider the following:

1. **OR**: Two heads are observed OR three heads are observed. 
2. **AND**: Not two tails are observed AND not three tails are observed.
3. **NOT**: Not more than one tails are observed.
::::
:::

We define mathematical operations to encapsulate the use of **or**, **and**, and **not**. These operations apply to any mathematical sets, whether they refer to events or not. 

:::{#def-union}
## Union
The union encodes the use of "or" in reference to two or more sets. Formally, with two sets $A$ and $B$, the union of $A$ and $B$ is the set of all elements that are contained in $A$, or $B$, or both $A$ and $B$. We write $A \cup B$ and read that as $A$ union $B$. When we wish to take the union of many sets, $A_1,A_2,\dots,A_n$, we write this as $$A_1 \cup A_2 \cup \cdots \cup A_n = \bigcup_{i=1}^n A_i.$$
:::

:::{#def-intersection}
## Intersection
The intersection captures the use of "and" in reference to two or more sets. Formally, the intersection of two sets, $A$, and $B$, is the set that contains all elements that belong to both $A$ and $B$. We write $A \cap B$, and say "$A$ intersect $B$." When we wish to take the intersection of many sets, $A_1,A_2,\dots,A_n$, we write $$A_1 \cap A_2 \cap \cdots \cap A_n = \bigcap_{i=1}^n A_i.$$
:::

:::{#def-complement}
## Complement
The complement makes formal the concept of "not." The complement of a set is the set of all elements which occur in the sample space but are not in the given set. We write this as $A^C$ and say "$A$ complement." When dealing with a sample space, $\mathcal{S}$, the complement of $A$ is the set of all elements in $\mathcal{S}$ that are not in $A$.
:::

:::{#exm-basic-set-operations}
## Basic Set Operations
For the game being played by Charles and Sadie, take $E_1 = \{(\text{H}, \text{H}, \text{H})\}$, $E_2 = \{(\text{H}, \text{H}, \text{H}), (\text{T}, \text{H}, \text{H}), (\text{H}, \text{T}, \text{H})\}$, and $E_3 = \{(\text{H}, \text{H}, \text{T})\}$. Express the following events.

a. $E_1 \cup E_2$;
b. $E_1 \cap E_2$;
c. $E_2^C$;
d. $E_2 \cap E_3$;
e. $E_1 \cup E_2 \cup E_3$.

::::{.callout .solution collapse='true'}
## Solution
Directly from definitions we can write down each of the following sets:


a. $$E_1 \cup E_2 =  \{(\text{H}, \text{H}, \text{H}), (\text{T}, \text{H}, \text{H}), (\text{H}, \text{T}, \text{H})\} = E_2.$$ As a result, the union of $E_1$ and $E_2$ is simply $E_2$.^[Note that whenever we have two events, $A$ and $B$, with $A\subseteq B$, then $A\cup B = B$.]
b. $$E_1 \cap E_2 = \{(\text{H}, \text{H}, \text{H})\} = E_1.$$ As a result, the intersection of $E_1$ and $E_2$ is simply $E_1$.^[Note that whenever we have two events, $A$ and $B$, with $A\subseteq B$ then $A \cap B = A$.]
c. $$E_2^C =  \{(\text{H},\text{H},\text{T}), (\text{H},\text{T},\text{T}), (\text{T},\text{H},\text{T}), (\text{T},\text{T},\text{H}), (\text{T},\text{T},\text{T})\}.$$
d. For $E_2\cap E_3$ note that they share no elements. As a result, the intersection will be empty since there are no elements common to both of them. This gives $E_2\cap E_3 = \emptyset$.
e. $$E_1 \cup E_2 \cup E_3 = \{(\text{H}, \text{H}, \text{H}), (\text{T}, \text{H}, \text{H}), (\text{H}, \text{T}, \text{H}), (\text{H}, \text{H}, \text{T})\}.$$
::::
:::

:::{#def-disjoint-events}
## Disjoint Events
Two events, $E_1$ and $E_2$ are said to be disjoint whenever their intersection is the null event. That is, if $E_1 \cap E_2 = \emptyset$ then $E_1$ and $E_2$ are disjoint events.
:::

These concepts allow us to more compactly express sets of interest, and in particular, will be quite useful when it comes to assigning probability. The more times you work with the set operations, the more familiar they will become, and as a result, practice is always useful. 

Consider rolling a 6-sided die, and take $A$ to be the event that a $6$ is rolled, $B$ to be the event that the roll was at least $5$, $C$ to be the event that the roll was less than $4$, and $D$ to be the event that the roll was odd.

* If we consider $D^C$ this is the event that the roll was even; 
* $A \cup C$ is the event that a $6$ was rolled or that a number less than $4$ was rolled, which is to say anything other than a $4$ or a $5$, which we may also express as $\{4,5\}^C$. 
* If we take  $A \cup B$ then this will be the same as $B$, and $A\cap B$ will be $A$. 
* If we take the event $A\cap C$, notice that no outcomes satisfy both conditions, and so $A \cap C = \emptyset$. 
* We can also join together multiple operations. $D^C \cap C$ gives us even numbers than less than $4$, which is to say the outcome $2$. 
* Similarly, $(A \cap B)^C$ would represent the event that a number less than $6$ is rolled. 

:::{#exm-card-set-operations}
## Set Operations with Decks of Cards
Charles and Sadie are tiring of flipping their coin, and so they wish to start using decks of cards sometimes instead. Before they formalize a game based on decks of cards, they want to make sure that they are both very comfortable working with these. Suppose that the sample space is defined to be the set of $52$ standard cards that may be drawn on a single draw from the deck. Describe how set operations can be used to form events corresponding to:

a. A red card is observed.
b. Any card between an ace and a ten is observed.
c. The ace of spades is observed.

::::{.callout .solution collapse='true'}
## Solution
First, we define several events. Note, these can be defined in shorthand to prevent needing to write out many different cards. we take $D$ to be the event that a diamond is observed, we take $H$ to be the event that a heart is observed, take $S$ to be the event that a spade is observed.^[Note, these three are compound events with $13$ different outcomes contained within them.] Take $A$ to be the event that an ace is observed, $J$ to be the event that a Jack is observed, $Q$ to be the event that a Queen is observed, and $K$ to be the event that a King is observed.^[These are all compound events with four different options.] We can use unions, intersections, and complements to express the previously mentioned scenarios. 

a. To represent outcomes corresponding to "the card is red", we can use $D \cup H$. 
b. To represent outcomes corresponding to "an ace through ten", we can use $(J \cup Q \cup K)^C$. 
c. To represent the outcome "the ace of spades", we may use $A \cap S$.
::::
:::

Working with these basic set operations should eventually become second nature. There are often very many ways of expressing the same event using these different operations, and finding the most useful method of representing a particular event can often be the key to solving challenging probability questions. The first step in making sure that these tools are available to you is in ensuring that the basic operations are fully understood, and this comes via practice. Remember, unions represent "ors", intersections represent "ands", and complements represent "nots". 

### Using R To Represent Sample Spaces and Events and Performing Set Operations
We have seen how R can encode sets of elements using vectors. For instance, we may take `sample_space <- 1:6` to represent the sample space of rolling a six-sided die. We can form events by taking subsets of the relevant quantities, selecting via indices. Fortunately, there are also all of the basic set operations implemented in R. 

We can use `union(x, y)` to perform the union of `x` and `y`, `intersect(x, y)` to perform the intersection of `x` and `y`, and `setdiff(x = sample_space, y)` to perform the complement of `y` (assuming that `sample_space` contains the full sample space).^[R does not implement complements directly, and instead implements the set difference operation. The set difference function, `setdiff(x, y)` returns the set of all elements in `x` which are not in `y`, a sort of subtracting of sets. The complement of a set is defined to be $A^C = \text{setdiff}(\mathcal{S}, A)$, indicating why this works!]

::: {.content-visible when-format='pdf'}
```{r}
# Define the Sample Space of Rolling a 20 Sided Die
sample_space <- 1:20

# Define some Events
E1 <- sample_space[2]
E2 <- sample_space[c(1, 3, 5, 7)]
E3 <- sample_space[c(1, 2, 4, 8)]

# Consider Set Operations
union(x = E1, y = E2) # E1 union E2 = {1, 2, 3, 5, 7}
union(x = E1, y = E3) # E1 union E3 = {1, 2, 4, 8}

intersect(x = E1, y = E3) # E1 intersect E3 = {2}
intersect(x = E1, y = E2) # E1 intersect E2 = {}

setdiff(x = sample_space, y = E1) # E1 complement

# (E2 union E3) complement
setdiff(x = sample_space, y = union(E2, E3))
```
:::
::: {.content-visible when-format='html'}
```{webr-r}
# Define the Sample Space of Rolling a 20 Sided Die
sample_space <- 1:20

# Define some Events
E1 <- sample_space[2]
E2 <- sample_space[c(1, 3, 5, 7)]
E3 <- sample_space[c(1, 2, 4, 8)]

# Consider Set Operations
union(x = E1, y = E2) # E1 union E2 = {1, 2, 3, 5, 7}
union(x = E1, y = E3) # E1 union E3 = {1, 2, 4, 8}

intersect(x = E1, y = E3) # E1 intersect E3 = {2}
intersect(x = E1, y = E2) # E1 intersect E2 = {}

setdiff(x = sample_space, y = E1) # E1 complement

# (E2 union E3) complement
setdiff(x = sample_space, y = union(E2, E3)) 
```
:::

## Venn Diagrams
The sample space is partitioned into outcomes, and the outcomes can be grouped together into events. These events are sets  and can be manipulated via basic set operations. Sometimes it is convenient to represent this process graphically through the use of Venn diagrams. In a Venn diagram, the sample space is represented by a rectangle with the possible outcomes placed inside, and events are drawn  inside of this as circles containing the relevant outcomes.

![A basic Venn diagram, representing the sample space and two different events. In practice, the sample space would have the possible outcomes written into the rectangle, and the circled events would end up containing the relevant outcomes for those events.](/graphics/ch2-venn-diagram-basic){#fig-venn-diagram}

On the Venn diagram then, the overlap between circles represents their intersection, the combined area of two (or more) circles represents their union, and everything outside of a given circle represents the complement. This can be a fairly useful method for representing sample spaces, and for visualizing the basic set operations that we use to manipulate events inside the sample spaces. 

A word of caution: Venn diagrams are useful tools, but they are not suitable as mathematical proofs directly. It is possible to convince yourself of false truths if the wrong diagrams are used, and as a result, Venn diagrams should be thought of as aids to understanding, rather than as a rigorous tool in and of themselves.^[This is a general principle in mathematics. Coming up with one example that makes something *seem* true does not form an argument demonstrating that it *is* true. Venn Diagrams should largely be thought of as specific examples of the underlying phenomena, which are great if you're a visual learner!]

![**Union:** The union of events A and B is shaded here in red. The union of two sets is all of the contents of both sets, including the overlap between the two.](/graphics/ch2-venn-diagram-union){#fig-venn-diagram-union}

![**Intersection:** The intersection of events A and B is shaded here in red. The intersection of two sets is all of the content shared by both sets, given by the overlapping area of the two circles.](/graphics/ch2-venn-diagram-intersection){#fig-venn-diagram-intersection}

![**Complement:** The complement of event A is shaded here in red. The complement of a sets is all of area inside of the sample space, not inside of the set. Here we show the complement of Event A, though Event B would be similar.](/graphics/ch2-venn-diagram-complement){#fig-venn-diagram-complement}

{{< pagebreak >}}

:::{#exm-charles-and-sadie-vd}
## Venn Diagram with Defined Events
Draw a Venn diagram representing the original game that Charles and Sadie played. On the diagram draw the events corresponding to "At least one head **and** one tail are observed", and "Sadie won the game". Recall that three coins are tossed, and Sadie wins if at least two of them show heads.

::::{.callout .solution collapse='true'}
## Solution
The sample space contains the eight possible options. Only $(\text{T}, \text{T}, \text{T})$ does not belong to at least one of the events. Both events share $(\text{H},\text{H},\text{T})$, $(\text{H},\text{T},\text{H})$, and $(\text{T},\text{H},\text{H})$. 

![](/graphics/ch2-venn-diagram-example)
::::
:::

Sample spaces, events, and the manipulation of these quantities forms a critical component of understanding probability models. In particular, they describe the complete set of occurrences in a statistical experiment that we could be interested in assigning probability values to. To formalize a probability model, however, we also need some rule for assigning probability values. 

## Exercises{.unnumbered}

:::{#exr-2.1}
For each of the following experiments, describe the relevant sample space and identify one possible event of interest.

a. The quality control inspection of smartphone screens from a manufacturing process.
a. Monitoring the ongoing structural integrity of a newly built bridge.
a. A clinical trial studying the effectiveness of a new drug.
a. Epidemiological monitoring of a disease outbreak.
a. Dealing a hand of black jack.
a. Observing the launch conditions for a rocket launch.
a. Debugging in software development.
a. Playing the lottery.
:::


:::{#exr-2.2}
A card is drawn at random from an ordinary deck of $52$ playing cards. Let $A$ be the event that a king is drawn, and $B$ the event that a club is drawn. In words, describe the following events.

a. $A\cup B$;
a. $A \cap B$;
a. $A\cup B^C$;
a. $A^C\cup B^C$;
a. $(A\cap B)\cup(A\cap B^C)$.
:::

:::{#exr-2.3}
Suppose that $\mathcal{S} = \{\phi, \lambda, \Delta, \mu\}$. List all possible events from the corresponding experiment.
:::

:::{#exr-2.4}
Suppose an experiment is run which generates realizations from positive integers.  Take $A$ to be the event $\{1, 5, 31, 56, 101\}$, $B = \{22, 56, 5, 103, 87\}$, $C = \{41, 13, 7, 101, 48\}$, and $D$ to be the event that the number is odd. Identify (write down or describe) each of the following events.

a. $D^C$
a. $A\cap B$
a. $C \cup A$
a. $C \cap D$
a. $(A\cup B)\cup (C\cup D)$
a. $A\cap D^C$
:::

:::{#exr-2.5}
Suppose that a $20$ sided die is rolled. The events of interest are: $A$ the outcome is a multiple of $4$, and $B$ the outcome is a multiple of $5$. 

a. Draw a Venn Diagram representing the sample space and events.
a. Identify the event $A \cup B$. What does this correspond to in words?
a. Identify the event $A \cap B^C$. What does this correspond to in words?
a. How would you denote the event ``neither a multiple of $4$ nor $5$.'' in terms of $A$ and $B$?
:::

:::{#exr-2.6}
Suppose that two indistinguishable coins are flipped. The events of interest are: $A$ exactly two heads are seen, and $B$ at least one head is seen.

a. Draw a Venn Diagram representing the sample space and events.
a. Describe every possible outcome with respect to the identified events.
a. Give an event, in terms of the number of heads observed, which is equivalent to the sample space.
:::

:::{#exr-2.7}
A cinema has $12$ screens, numbered $1$ through $12$. Before opening, an employee checks to ensure that the projectors are correctly calibrated. 

Let $A$ be the event that all the screens are correctly calibrated, $B$ be the event that the third screen is not correctly calibrated, $C$ be the event that exactly one screen is not correctly calibrated, and $D$ be the event that $5$ and $8$ are correctly calibrated.

Which of the following pairs of events are disjoint?

a. $A$ and $B$.
a. $B$ and $D$. 
a. $C$ and $D$.
a. $B$ and $C$.
:::


::: {.content-visible when-format='html'}

## Self-Assessment {.unnumbered}

Note: the following questions are still experimental. Please contact me if you have any issues with these components. This can be if there are incorrect answers, or if there are any technical concerns. Each question currently has an ID with it, randomized for each version. If you have issues, reporting the specific ID will allow for easier checking!

For each question, you can check your answer using the checkmark button. You can cycle through variants of the question by pressing the arrow icon. 


```{r}
#| echo: false
#| message: false
#| warning: false

library(exams2forms)
set.seed(31415)

```

:::{#sa-2.01}
```{r}
#| echo: false
#| message: false
#| results: 'asis'
#| warning: false
#| cache: true
exams2forms("2.A.EventDefinition.Rmd", 
            edir = "../PracticeQuestions/Chapter2", 
            n = 25)
```
:::

:::{#sa-2.02}
```{r}
#| echo: false
#| message: false
#| results: 'asis'
#| warning: false
#| cache: true
cache_check <- 123456789
exams2forms("2.B.SampleSpaceDefinition.Rmd", 
            edir = "../PracticeQuestions/Chapter2", 
            n = 15)
```
:::

:::{#sa-2.03}
```{r}
#| echo: false
#| message: false
#| results: 'asis'
#| warning: false
#| cache: true
cache_check <- 12345678900

exams2forms("2.C.SimpleVersusCompoundEvents.Rmd", 
            edir = "../PracticeQuestions/Chapter2", 
            n = 50)
```
:::

:::{#sa-2.04}
```{r}
#| echo: false
#| message: false
#| results: 'asis'
#| warning: false
#| cache: true
cache_check <- 12345678900

exams2forms("2.D.BasicSetOperations.Rmd", 
            edir = "../PracticeQuestions/Chapter2", 
            n = 100)
```
:::

:::{#sa-2.05}
```{r}
#| echo: false
#| message: false
#| results: 'asis'
#| warning: false
#| cache: true
cache_check <- 123456789

exams2forms("2.E.DisjointEvents.Rmd", 
            edir = "../PracticeQuestions/Chapter2", 
            n = 20)
```
:::

:::