```{r data generation, echo = FALSE, results = "hide"}

increase <- c("The sample size, $n$, is decreased", 
             "The population variance is increased.",
             "The population standard deviation is increased.",
             "The confidence level is increased.",
             "The level of significance, $\\alpha$, is decreased.")

decrease <- c("The sample size, $n$, is increased", 
             "The population variance is decreased.",
             "The population standard deviation is decreased.",
             "The confidence level is decreased.",
             "The level of significance, $\\alpha$, is increased.")

no_change <- c(
    "The sample mean is increased.",
    "The sample mean is decreased.",
    "The population mean is increased.",
    "The population mean is decreased.",
    "The sample standard deviation is increased.",
    "The sample standard deviation is decreased."
)


inc <- sample(1:3, 1)
dec <- sample(1:(4-inc), 1)
nc <- 5 - inc - dec

inc_statements <- sample(increase, inc)
dec_statements <- sample(decrease, dec)
nc_statements <- sample(no_change, nc)


if(sample(0:1,1) == 0) {
    answers <- c(inc_statements, dec_statements, nc_statements)
    solutions <- c(rep(TRUE, length(inc_statements)), rep(FALSE, 5 - length(inc_statements)))
    type <- "increase"
} else {
    answers <- c(dec_statements, inc_statements, nc_statements)
    solutions <- c(rep(TRUE, length(dec_statements)), rep(FALSE, 5 - length(dec_statements)))
    type <- "decrease"
}

o <- sample(1:5)
answers <- answers[o]
solutions <- solutions[o]

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========
Consider a $100\times(1-\alpha)\%$ confidence interval for the mean of a normal distribution with known variance, $\sigma^2$, based on a random sample of $n$ observations.

For which of the following changes will the width of the confidence interval `r type`, if everything else is held constant?

(Question ID: *`r id`*)

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(answers, markup = "markdown")
```

Meta-information
================
extype: mchoice
exsolution: `r mchoice2string(solutions)`
exname: 14.10 - Factors Influencing CI Width