```{r data generation, echo = FALSE, results = "hide"}

paste_set <- function(S) {
  paste0("\\{", paste(S, collapse=", "), "\\}")
}

## Generate three Sets
s <- sample(0:10, 1)

S <- (s):(s + 9)
A <- sort(sample(S, 4, FALSE))
B <- sort(sample(S, 4, FALSE))
C <- sort(sample(S, 4, FALSE))

event_and_answer <- list(
    list(event = "$A^C$", ans = setdiff(S, A)),
    list(event = "$B^C$", ans = setdiff(S, B)),
    list(event = "$C^C$", ans = setdiff(S, C)),
    list(event = "$A\\cup B$", ans = union(A, B)),
    list(event = "$A\\cup C$", ans = union(A, C)),
    list(event = "$B\\cup C$", ans = union(B, C)),
    list(event = "$A\\cap B$", ans = intersect(A, B)),
    list(event = "$A\\cap C$", ans = intersect(A, C)),
    list(event = "$B\\cap C$", ans = intersect(B, C)),
    list(event = "$A\\cup B^C$", ans = union(A, setdiff(S, B))),
    list(event = "$A\\cup C^C$", ans = union(A, setdiff(S, C))),
    list(event = "$B\\cup C^C$", ans = union(B, setdiff(S, C))),
    list(event = "$A^C\\cup B$", ans = union(setdiff(S, A), B)),
    list(event = "$A^C\\cup C$", ans = union(setdiff(S, A), C)),
    list(event = "$B^C\\cup C$", ans = union(setdiff(S, B), C)),
    list(event = "$A\\cap B^C$", ans = intersect(A, setdiff(S, B))),
    list(event = "$A\\cap C^C$", ans = intersect(A, setdiff(S, C))),
    list(event = "$B\\cap C^C$", ans = intersect(B, setdiff(S, C))),
    list(event = "$A^C\\cap B$", ans = intersect(setdiff(S, A), B)),
    list(event = "$A^C\\cap C$", ans = intersect(setdiff(S, A), C)),
    list(event = "$B^C\\cap C$", ans = intersect(setdiff(S, B), C))
)

which_q <- sample(1:length(event_and_answer), 1)
true_set <- event_and_answer[[which_q]]$ans
event_of_interest <- event_and_answer[[which_q]]$event 
solutions <- S %in% true_set

id <- sprintf("%010d", sample(0:1000000000, 1))
```

Question
========
Suppose that the sample space for an experiment, $\mathcal{S}$ is given by $`r paste_set(S)`$. 
Moreover, take three events, $A = `r paste_set(A)`$, $B = `r paste_set(B)`$, and $C = `r paste_set(C)`$.

Which of the following elements comprise `r event_of_interest`? (Note, select all that apply).

(Question ID: *`r id`*)

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(S, markup = "markdown")
```

Meta-information
================
extype: mchoice
exsolution: `r mchoice2string(solutions, single = FALSE)`
exname: 2.04 - Basic set operations.