```{r data generation, echo = FALSE, results = "hide"}

library(gtsummary)
library(ggplot2)
library(MASS)
get_article_num <- function(num) {
    fd <- as.numeric(substr(num, 1, 1))

    if(fd == 8) {
        "an"
    } else {
        "a"
    }
}

scenarios <- list(
  list(
    scenario = "A store owner wants to use linear regression to understand the relationship between a diamond's cost and the number of carats it is.",
    outcome = "cost",
    predictor = "carat",
    intercept = -2250,
    slope = 7750,
    minx = 0.2,
    maxx = 5.1,
    eps_sd = 1500
  ),
  list(
    scenario = "A biologist is investigating the relationship between the height of a tree and its age.",
    outcome = "height",
    predictor = "age",
    intercept = 2,
    slope = 0.5,
    minx = 5,
    maxx = 100,
    eps_sd = 1
  ),
  list(
    scenario = "A researcher is studying if there is a link between the amount of fertilizer used and the yield of corn.",
    outcome = "corn yield",
    predictor = "fertilizer amount",
    intercept = 50,
    slope = 0.8,
    minx = 0,
    maxx = 200,
    eps_sd = 5
  ),
  list(
    scenario = "An economist wants to see if there's a relationship between a country's GDP and its average life expectancy.",
    outcome = "life expectancy",
    predictor = "GDP",
    intercept = 65,
    slope = 0.0001,
    minx = 1000,
    maxx = 100000,
    eps_sd = 2
  ),
  list(
    scenario = "A teacher is examining if the number of hours studied affects a student's exam score.",
    outcome = "exam score",
    predictor = "study hours",
    intercept = 40,
    slope = 7,
    minx = 0,
    maxx = 15,
    eps_sd = 8
  ),
  list(
    scenario = "A marketing team is exploring the relationship between the amount spent on advertising and product sales.",
    outcome = "product sales",
    predictor = "advertising spend",
    intercept = 100,
    slope = 2.5,
    minx = 50,
    maxx = 500,
    eps_sd = 20
  ),
  list(
    scenario = "A scientist is investigating if there is a relationship between the temperature of a reaction and its rate.",
    outcome = "reaction rate",
    predictor = "temperature",
    intercept = 0.1,
    slope = 0.02,
    minx = 20,
    maxx = 80,
    eps_sd = 0.05
  ),
  list(
    scenario = "A farmer is looking at the relationship between rainfall and crop yield.",
    outcome = "crop yield",
    predictor = "rainfall",
    intercept = 30,
    slope = 0.6,
    minx = 10,
    maxx = 100,
    eps_sd = 7
  ),
  list(
    scenario = "A doctor is studying the association between a patient's age and their cholesterol level.",
    outcome = "cholesterol level",
    predictor = "age",
    intercept = 150,
    slope = 1.2,
    minx = 20,
    maxx = 80,
    eps_sd = 15
  ),
  list(
    scenario = "A software engineer is analyzing if the number of lines of code affects the number of bugs.",
    outcome = "number of bugs",
    predictor = "lines of code",
    intercept = 2,
    slope = 0.01,
    minx = 100,
    maxx = 5000,
    eps_sd = 3
  ),
  list(
    scenario = "A city planner is examining the relationship between the distance from the city center and housing price.",
    outcome = "housing price",
    predictor = "distance from city center",
    intercept = 500000,
    slope = -5000,
    minx = 1,
    maxx = 20,
    eps_sd = 30000
  ),
  list(
    scenario = "A car manufacturer is looking at the relationship between engine size and fuel efficiency.",
    outcome = "fuel efficiency",
    predictor = "engine size",
    intercept = 40,
    slope = -5,
    minx = 1.0,
    maxx = 4.0,
    eps_sd = 3
  ),
  list(
    scenario = "A musician is exploring if there is a relationship between the number of practice hours and performance quality.",
    outcome = "performance quality",
    predictor = "practice hours",
    intercept = 3,
    slope = 0.8,
    minx = 0,
    maxx = 20,
    eps_sd = 1.5
  ),
  list(
    scenario = "A student is investigating if there is a relationship between the number of siblings and their grade point average.",
    outcome = "grade point average",
    predictor = "number of siblings",
    intercept = 3.0,
    slope = 0,
    minx = 0,
    maxx = 5,
    eps_sd = 0.4
  ),
  list(
    scenario = "A meteorologist is examining the relationship between air pressure and the probability of rain.",
    outcome = "probability of rain",
    predictor = "air pressure",
    intercept = 0.8,
    slope = -0.0005,
    minx = 980,
    maxx = 1030,
    eps_sd = 0.1
  ),
  list(
    scenario = "A fitness instructor is looking at the relationship between the number of push-ups a person can do and their cardiovascular health score.",
    outcome = "cardiovascular health score",
    predictor = "number of push-ups",
    intercept = 60,
    slope = 0.5,
    minx = 0,
    maxx = 50,
    eps_sd = 5
  ),
  list(
    scenario = "A coffee shop owner wants to see if there's a relationship between the temperature of a latte and customer satisfaction.",
    outcome = "customer satisfaction",
    predictor = "latte temperature",
    intercept = 5,
    slope = 0,
    minx = 140,
    maxx = 170,
    eps_sd = 1.2
  ),
  list(
    scenario = "A social scientist is studying the relationship between the number of social media followers and perceived popularity.",
    outcome = "perceived popularity",
    predictor = "number of followers",
    intercept = 2,
    slope = 0.00001,
    minx = 100,
    maxx = 100000,
    eps_sd = 0.8
  )
)

pretty_round <- function(num, digits) {
    if(abs(num) > 10^(digits - 1)) {
        round(num, 0)
    } else {
        signif(num, digits)
    }
}

sc_idx <- sample(1:length(scenarios), 1)

n <- sample(30:100, 1)
scenario <- scenarios[[sc_idx]]$scenario 
pred_name <- scenarios[[sc_idx]]$predictor
outcome_name <- scenarios[[sc_idx]]$outcom

beta0 <- scenarios[[sc_idx]]$intercept
beta1 <- scenarios[[sc_idx]]$slope
X <- runif(n, scenarios[[sc_idx]]$minx, scenarios[[sc_idx]]$maxx)
Y <- beta0 + beta1 * X + rnorm(n, 0, scenarios[[sc_idx]]$eps_sd)

fitted_model <- lm(Y ~ X)
names(fitted_model$coefficients) <- c('Intercept', stringr::str_to_title(pred_name))

xc <- round(runif(1, min(X), max(X)), 2)
xp <- round(runif(1, min(X), max(X)), 2)

desired_test_stat <- round(runif(1, -4, 4), 2)
seb0 <- pretty_round(summary(fitted_model)$coefficients[1,2], 2)
seb1 <- pretty_round(summary(fitted_model)$coefficients[2,2], 2)
estb0 <- pretty_round(fitted_model$coefficients[1], 2)
estb1 <- pretty_round(fitted_model$coefficients[2], 2)

null_val <- round(estb1 - desired_test_stat * seb1, 2)

clp <- sample(80:99, 1)

xbar <- round(mean(X), 2)
sd_x <- round(sd(X), 2)

crit_v <- abs(qt(0.5 - clp/200, df = n - 2))

sigma_v <- pretty_round(sigma(fitted_model), 3)

ans1 <- estb0 + estb1 * xc
ans2 <- (estb1 - null_val)/seb1
ans3 <- estb0 - seb0 * crit_v
ans4 <- estb0 + seb0 * crit_v
ans5 <- estb1 - seb1 * crit_v
ans6 <- estb1 + seb1 * crit_v
ans7 <- estb0 + estb1 * xp - crit_v * sigma_v * sqrt(1/n + (xp - xbar)^2/((n-1)*sd_x^2))
ans8 <- estb0 + estb1 * xp + crit_v * sigma_v * sqrt(1/n + (xp - xbar)^2/((n-1)*sd_x^2))

ans9 <- estb0 + estb1 * xp - crit_v * sigma_v * sqrt(1 + 1/n + (xp - xbar)^2/((n-1)*sd_x^2))
ans10 <- estb0 + estb1 * xp + crit_v * sigma_v * sqrt(1 + 1/n + (xp - xbar)^2/((n-1)*sd_x^2))
id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========

`r scenario` 

The results of the analysis are summarized in the following regression table. 

```{r}
#| echo: false
#| warning: false
#| output: asis

fitted_model |> 
    tbl_regression(intercept = TRUE) |> 
    modify_column_hide(column = c(conf.low)) |>
    modify_column_unhide(column = std.error) |>
    add_glance_source_note(include = c(nobs, sigma, r.squared))
```

The mean `r pred_name` is `r xbar`, with a standard deviation of `r sd_x`. 

a. What is the predicted `r outcome_name` if `r pred_name` is `r xc`? ##ANSWER1##
b. What is the value of the test statistic for testing $H_0: \beta_1 = `r null_val`$? ##ANSWER2##
c. Give `r get_article_num(clp)` $`r clp`\%$ confidence interval for $\beta_0$.

Lower Bound: ##ANSWER3## Upper Bound: ##ANSWER4##

d. Give `r get_article_num(clp)` $`r clp`\%$ confidence interval for $\beta_1$

Lower Bound: ##ANSWER5## Upper Bound: ##ANSWER6##

e. Give `r get_article_num(clp)` $`r clp`\%$ confidence interval for the expected `r outcome_name`, if the `r pred_name` is `r xp`. 

Lower Bound: ##ANSWER7## Upper Bound: ##ANSWER8##

f. Give `r get_article_num(clp)` $`r clp`\%$ prediction interval for the `r outcome_name`, if the `r pred_name` is `r xp`. 

Lower Bound: ##ANSWER9## Upper Bound: ##ANSWER10##


(Question ID: *`r id`*)

Meta-information
================
extype: cloze
exclozetype: num|num|num|num|num|num|num|num|num|num
exsolution: `r fmt(ans1, 3)`|`r fmt(ans2, 3)`|`r fmt(ans3, 3)`|`r fmt(ans4, 3)`|`r fmt(ans5, 3)`|`r fmt(ans6, 3)`|`r fmt(ans7, 3)`|`r fmt(ans8, 3)`|`r fmt(ans9, 3)`|`r fmt(ans10, 3)`
extol: 0.01
exname: Regression questions from regression table