```{r data generation, echo = FALSE, results = "hide"}

sided <- sample(1:3, 1)
sc <- sample(1:8, 1)

# Question Texts 


if(sc == 1) {
    qscenario <- "A group of researchers are interested in studying the poverty rates for two different regions."
    pop1 <- "individuals living in the first region"
    pop2 <- "individuals living in the second region"

    directions <- c("experience higher rates of poverty", 
                    "experience lower rates of poverty", 
                    "experience different rates of poverty")

    pX <- runif(1, 0.05, 0.15)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 2) {
    qscenario <- "A group of researchers are interested in studying the rates of smartphone ownership for students from two different schoolboards."
    pop1 <- "students from the first school board"
    pop2 <- "students from the second school board"

    directions <- c("have a higher rate of smartphone ownership", 
                    "have a lower rate of smartphone ownership", 
                    "have different rates of smartphone ownership")

    pX <- runif(1, 0.4, 0.85)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 3) {
    qscenario <- "Researchers are considering the proportion of individuals from two different countries who smoke cigarettes."
    pop1 <- "residents of the first country"
    pop2 <- "residents of the second country"

    directions <- c("smoke in higher proportion", 
                    "smoke in lower proportion", 
                    "smoke in a different proportion")

    pX <- runif(1, 0.05, 0.45)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 4) {
    qscenario <- "A researcher is interested in determining whether male and female children exhibit left-hand dominance with different rates."
    pop1 <- "male children"
    pop2 <- "female children"


    directions <- c("are left-handed in a higher proportion of cases", 
                    "are left-handed in a lower proportion of cases", 
                    "are left-handed in a different proportion of cases")

    pX <- runif(1, 0.05, 0.15)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 5) {
    qscenario <- "A municipality is interested in determining whether residents of regions within the municipality recycle at different rates."
    pop1 <- "residents of the first region"
    pop2 <- "residents of the second region"

    directions <- c("have a higher rate of recycling", 
                    "have a lower rate of recycling", 
                    "have a different rate of recycling")

    pX <- runif(1, 0.1, 0.9)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 6) {
    qscenario <- "A researcher is interested in determining whether there is a difference in the rates of veganism between females and males."
    pop1 <- "females"
    pop2 <- "males"

    directions <- c("have a higher probability of being vegan", 
                    "have a lower probability of being vegan ", 
                    "have a different probability of being vegan")

    pX <- runif(1, 0.02, 0.15)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 7) {
    qscenario <- "A sociologist is studying the rates of daily social media usage, comparing young and older people."
    pop1 <- "young people"
    pop2 <- "older people"

    directions <- c("have a higher rate of daily social media usage", 
                    "have a lower rate of daily social media usage", 
                    "have a different rate of daily social media usage")

    pX <- runif(1, 0.35, 0.9)
    n <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)

    m <- sample(1000:2000, 1)

} else if(sc == 8) {
    qscenario <- "An organizational behaviour specialist is interested in whether the rates of job satisfaction differ between people working in human resources and those working in finance."
    pop1 <- "people working in human resources"
    pop2 <- "people working in finance"

    directions <- c("report higher rates of being satisfied with their job", 
                    "report lower rates of being satisfied with their job", 
                    "report different rates of job satisfaction")

    pX <- runif(1, 0.3, 0.9)
    n <- sample(1000:2000, 1)
    m <- sample(1000:2000, 1)
    X <- rbinom(1, n, pX)


} 


direction <- directions[sided]
# Sided = 1: X > Y
# Sided = 2: X < Y
# Sided = 3: X = Y

conclusion <- sample(0:1, 1)
alpha <- round(runif(1, 0.01, 0.2), 2)

# Sample a p-value based on the desired conclusion and alpha
if(conclusion == 0) {
    # Reject H0
    p <- round(runif(1, 0.001, alpha - 0.005), 3)
} else {
    # Fail to Reject H0
    p <- round(runif(1, alpha + 0.001, 0.9), 3)
}

# Compute the Relevant values for Ybar
if(sided == 1) {
    # One-Tail Upper
    crit <- qnorm(1-p)
    p_calc <- function(x) { 1 - pnorm(x) }

    hypotheses <- "$H_0: p_X \\leq p_Y$ vs. $H_A: p_X > p_Y$"

} else if(sided == 2) {
    # One-Tail Lower
    crit <- qnorm(p)
    p_calc <- function(x) { pnorm(x) }

    hypotheses <- "$H_0: p_X \\geq p_Y$ vs. $H_A: p_X < p_Y$"
} else {
    # Two-Tailed
    crit <- qnorm(sample(c(p/2, 1-p/2), 1))
    p_calc <- function(x) { 2*pnorm(-abs(x)) }

    hypotheses <- "$H_0: p_X = p_Y$ vs. $H_A: p_X \\neq p_Y$"
}


find_y <- function(yhat) {
    (X/n - yhat/m) / sqrt((X + yhat)/(n+m) * (1 - (X + yhat)/(n+m)) * (1/n + 1/m)) - crit
}
Y <- round(uniroot(f = find_y, interval = c(0, m))$root)

test_stat <- (X/n - Y/m) / sqrt((X + Y)/(n+m) * (1 - (X + Y)/(n+m)) * (1/n + 1/m))
p_value <- p_calc(test_stat)

###########################################################################
# Fixed Answers
all_hypotheses <- c(
    "$H_0: p_X \\leq p_Y$ vs. $H_A: p_X > p_Y$",
    "$H_0: p_X \\geq p_Y$ vs. $H_A: p_X < p_Y$",
    "$H_0: p_X = p_Y$ vs. $H_A: p_X \\neq p_Y$",
    "$H_0: p_X < p_Y$ vs. $H_A: p_X \\geq p_Y$",
    "$H_0: p_X > p_Y$ vs. $H_A: p_X \\leq p_Y$",
    "$H_0: p_X \\neq p_Y$ vs. $H_A: p_X = p_Y$"
)
answer_a <- c(TRUE, rep(FALSE, 4))
options_a <- c(hypotheses, sample(setdiff(all_hypotheses, hypotheses), 4))
sort_a <- sample(1:5, 5)

incorrect_conclusions <- c(
    "Because the $p$-value is greater than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we fail to reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we accept the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we accept the null hypothesis."
)
correct_conclusion <- ifelse(p_value <= alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)
incorrect_conclusion_include <- ifelse(p_value > alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)

answer_d <- c(TRUE, rep(FALSE, 4))
options_d <- c(correct_conclusion, incorrect_conclusion_include, sample(incorrect_conclusions, 3))
sort_d <- sample(1:5, 5)

ans1 <- answer_a[sort_a]
ans2 <- test_stat
ans3 <- p_value
ans4 <- answer_d[sort_d]

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========

`r qscenario` They collect a sample of `r n` `r pop1` (labeled $X$). Further, they take a sample of `r m` `r pop2` (labeled $Y$). 

From their sample, they observe $X = `r X`$ and $Y = `r Y`$. 

The interest of the study is in testing whether `r pop1` `r direction` (compared to `r pop2`).

a. What set of hypotheses should be tested in this setting? 

##ANSWER1##

b. What is the value of the test statistic used to test the hypothesis identified in (a)? ##ANSWER2##
c. What is the $p$-value computed for the hypothesis test? ##ANSWER3##
d. What is the conclusion from the hypothesis test, at a significance level of `r alpha`.

##ANSWER4##

Question ID: *`r id`*

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(c(options_a[sort_a], options_d[sort_d]), markup = "markdown")
```


Meta-information
================
extype: cloze
exsolution: `r mchoice2string(ans1)`|`r fmt(ans2, 3)`|`r fmt(ans3, 3)`|`r mchoice2string(ans4)`
exclozetype: schoice|num|num|schoice
exname: TwoSampleZTest
extol: 0.01