```{r data generation, echo = FALSE, results = "hide"}

sided <- sample(1:3, 1)
sc <- sample(1:11, 1)

# Question Texts 

if(sc == 1) {
    qscenario <- "A group of researchers are studying the impact of two different medications on blood pressure."
    pop1 <- "individuals taking the first medication"
    pop2 <- "individuals taking the second medication"

    directions <- c("have higher average systolic blood pressures", 
                    "have lower average systolic blood pressures", 
                    "have different average systolic blood pressures")

    sigmaX <- round(runif(1, 7, 13), 2)
    sigmaY <- round(runif(1, 7, 13), 2)
    Xbar <- round(runif(1, 100, 130), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 2) {
    qscenario <- "A group of researchers are studying the heights of individuals from two different regions."
    pop1 <- "individuals from the first region"
    pop2 <- "individuals from the second region"

    directions <- c("are taller on average", 
                    "are shorter on average", 
                    "are different height on average")

    sigmaX <- round(runif(1, 3, 10), 2)
    sigmaY <- round(runif(1, 3, 10), 2)
    Xbar <- round(runif(1, 155, 180), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 3) {
    qscenario <- "A medical researcher in interested in understanding the impact of two different diets on blood pressure."
    pop1 <- "individuals on the first diet"
    pop2 <- "individuals on the second diet"

    directions <- c("have higher average systolic blood pressures", 
                    "have lower average systolic blood pressures", 
                    "have different average systolic blood pressures")

    sigmaX <- round(runif(1, 7, 13), 2)
    sigmaY <- round(runif(1, 7, 13), 2)
    Xbar <- round(runif(1, 100, 130), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 4) {
    qscenario <- "A pyschologist is studying the reaction times of individuals who have taken two different medical treatments."
    pop1 <- "individuals taking the first treatment"
    pop2 <- "individuals taking the second treatment"


    directions <- c("have longer average reaction times", 
                    "have shorter average reaction times", 
                    "differ in their average reaction times")

    sigmaX <- round(runif(1, 20, 100), 2)
    sigmaY <- round(runif(1, 20, 100), 2)
    Xbar <- round(runif(1, 200, 800), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 5) {
    qscenario <- "A research nurse wants to understand how the presence (or absence) of prenatal vitamins impacts the weight of newborns."
    pop1 <- "newborns whose mothers took prenatal viatmins"
    pop2 <- "newborns whose mothers did not take prenatal vitamins"

    directions <- c("weigh more on average", 
                    "weigh less on average", 
                    "differ in their average weights")

    sigmaX <- round(runif(1, 200, 800), 2)
    sigmaY <- round(runif(1, 200, 800), 2)
    Xbar <- round(runif(1, 3000, 3800), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 6) {
    qscenario <- "Medical researchers are interested in studying how internal body temperatures are impacted by two different medications."
    pop1 <- "individuals taking the first medication"
    pop2 <- "individuals taking the second medication"

    directions <- c("have higher average body temperatures", 
                    "have lower average body temperatures", 
                    "have different average body temperatures")

    sigmaX <- round(runif(1, 0.5, 1.5), 4)
    sigmaY <- round(runif(1, 0.5, 1.5), 4)
    Xbar <- round(runif(1, 33, 38), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 7) {
    qscenario <- "Researchers are investigating the serum cholesterol levels of individuals who consume two different diets."
    pop1 <- "individuals on the first diet"
    pop2 <- "individuals on the second diet"

    directions <- c("have higher average serum cholesterol levels", 
                    "have lower average serum cholesterol levels", 
                    "have different average serum cholesterol levels")

    sigmaX <- round(runif(1, 15, 50), 2)
    sigmaY <- round(runif(1, 15, 50), 2)
    Xbar <- round(runif(1, 150, 220), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 8) {
    qscenario <- "Researchers are studying the red blood cell counts for individuals from two different populations."
    pop1 <- "individuals in the first population"
    pop2 <- "individuals in the second population"

    directions <- c("have higher average red blood cell counts", 
                    "have lower average red blood cell counts", 
                    "differ in their average red blood cell counts")

    sigmaX <- round(runif(1, 0.2, 0.8), 4)
    sigmaY <- round(runif(1, 0.2, 0.8), 4)
    Xbar <- round(runif(1, 3, 6.5), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 9) {
    qscenario <- "A lightbulb manufacturer wants to compare the lifespan for two different models of their lightbulbs."
    pop1 <- "lightbulbs of the first model"
    pop2 <- "lightbulbs of the second model"

    directions <- c("last longer on average", 
                    "last shorter on average", 
                    "last for a different average length of time")

    Xbar <- sample(2000:15000, 1)
    sigmaX <- round(runif(1, 50 * Xbar/2000, 100 * Xbar/2000), 2)
    sigmaY <- round(runif(1, 50 * Xbar/2000, 100 * Xbar/2000), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 10) {
    qscenario <- "A group of marine biologists wants to understand whether differences in length exist between male and female hickory shad."
    pop1 <- "male hickory shad"
    pop2 <- "female hickory shad"
    
    directions <- c("are longer on average", 
                    "are shorter on average", 
                    "have different average lengths")

    sigmaX <- round(runif(1, 2, 5), 2)
    sigmaY <- round(runif(1, 2, 5), 2)
    Xbar <- round(runif(1, 30, 40), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

} else if(sc == 11) {
    qscenario <- "A professor is interested in comparing the average number of hours that students study for Introductory Probability compared to Introductory Statistics."
    pop1 <- "students in Introductory Probability"
    pop2 <- "students in Introductory Statistics"

    directions <- c("study for more time", 
                    "study for less time", 
                    "study for different lengths of time")

    sigmaX <- round(runif(1, 5, 15), 2)
    sigmaY <- round(runif(1, 5, 15), 2)
    Xbar <- round(runif(1, 20, 50), 2)
    n <- sample(30:500, 1)
    m <- sample(30:500, 1)

}

direction <- directions[sided]
# Sided = 1: X > Y
# Sided = 2: X < Y
# Sided = 3: X = Y

conclusion <- sample(0:1, 1)
alpha <- round(runif(1, 0.01, 0.2), 2)

sp_squared <- ((n-1)*sigmaX^2 + (m-1)*sigmaY^2)/(n+m-2)

my_df <- (sigmaX^2/n + sigmaY^2/m)^2/((sigmaX^2/n)^2/(n-1) + (sigmaY^2/m)^2/(m-1))

# Sample a p-value based on the desired conclusion and alpha
if(conclusion == 0) {
    # Reject H0
    p <- round(runif(1, 0.001, alpha - 0.005), 3)
} else {
    # Fail to Reject H0
    p <- round(runif(1, alpha + 0.001, 0.9), 3)
}

# Compute the Relevant values for Ybar
if(sided == 1) {
    # One-Tail Upper
    crit <- qt(1-p, df = n+m-2)
    p_calc <- function(x) { 1 - pt(x, df = n+m-2) }

    hypotheses <- "$H_0: \\mu_X \\leq \\mu_Y$ vs. $H_A: \\mu_X > \\mu_Y$"

} else if(sided == 2) {
    # One-Tail Lower
    crit <- qt(p, df = n+m-2)
    p_calc <- function(x) { pt(x, df = n+m-2) }

    hypotheses <- "$H_0: \\mu_X \\geq \\mu_Y$ vs. $H_A: \\mu_X < \\mu_Y$"
} else {
    # Two-Tailed
    crit <- qt(sample(c(p/2, 1-p/2), 1), df = n+m-2)
    p_calc <- function(x) { 2*pt(-abs(x), df = n+m-2) }

    hypotheses <- "$H_0: \\mu_X = \\mu_Y$ vs. $H_A: \\mu_X \\neq \\mu_Y$"
}

# Find Ybar
Ybar <- round(Xbar - crit * sqrt(sp_squared* (1/ n + 1 / m)), 2)
test_stat <- (Xbar - Ybar)/sqrt(sp_squared* (1/ n + 1 / m))
p_value <- p_calc(test_stat)

###########################################################################
## Part (A)
all_hypotheses <- c(
    "$H_0: \\mu_X \\leq \\mu_Y$ vs. $H_A: \\mu_X > \\mu_Y$",
    "$H_0: \\mu_X \\geq \\mu_Y$ vs. $H_A: \\mu_X < \\mu_Y$",
    "$H_0: \\mu_X = \\mu_Y$ vs. $H_A: \\mu_X \\neq \\mu_Y$",
    "$H_0: \\mu_X < \\mu_Y$ vs. $H_A: \\mu_X \\geq \\mu_Y$",
    "$H_0: \\mu_X > \\mu_Y$ vs. $H_A: \\mu_X \\leq \\mu_Y$",
    "$H_0: \\mu_X \\neq \\mu_Y$ vs. $H_A: \\mu_X = \\mu_Y$"
)
answer_a <- c(TRUE, rep(FALSE, 4))
options_a <- c(hypotheses, sample(setdiff(all_hypotheses, hypotheses), 4))
sort_a <- sample(1:5, 5)

## Part (C)
answer_c <- c(TRUE, rep(FALSE, 4))
options_c <- c(
    paste0("A $t$ distribution with ", n + m -2, " degrees of freedom."),
    paste0("A $t$ distribution with ", round(my_df, 3), " degrees of freedom."),
    paste0("A $t$ distribution with ", n-1, " degrees of freedom."),
    paste0("A $t$ distribution with ", m-1, " degrees of freedom."),
    paste0("A standard normal distribution.")
)
sort_c <- sample(1:5, 5)

## Part (E)
incorrect_conclusions <- c(
    "Because the $p$-value is greater than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we fail to reject the null hypothesis.",
    "Because the $p$-value is smaller than $\\alpha$, we accept the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we accept the null hypothesis."
)
correct_conclusion <- ifelse(p_value <= alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)
incorrect_conclusion_include <- ifelse(p_value > alpha, 
    "Because the $p$-value is smaller than $\\alpha$, we reject the null hypothesis.",
    "Because the $p$-value is greater than $\\alpha$, we fail to reject the null hypothesis."
)

answer_e <- c(TRUE, rep(FALSE, 4))
options_e <- c(correct_conclusion, incorrect_conclusion_include, sample(incorrect_conclusions, 3))
sort_e <- sample(1:5, 5)

ans1 <- answer_a[sort_a]
ans2 <- test_stat
ans3 <- answer_c[sort_c]
ans4 <- p_value
ans5 <- answer_e[sort_e]

id <- sprintf("%010d", sample(0:1000000000, 1))

```

Question
========

`r qscenario` They collect a sample of `r n` `r pop1` (labeled $X$). Further, they take a sample of `r m` `r pop2` (labeled $Y$). 

From their sample, they observe $\overline{X} = `r format(Xbar, scientific=FALSE)`$ (with sample standard deviation $`r sigmaX`$) and $\overline{Y} = `r format(Ybar, scientific=FALSE)`$ (with sample standard deviation $`r sigmaY`$). From past research, there is good reason to suspect that the population variances for `r pop1` and `r pop2` are, in fact, equal. 

The interest of the study is in testing whether `r pop1` `r direction` (compared to `r pop2`).

a. What set of hypotheses should be tested in this setting? 

##ANSWER1##

b. What is the value of the test statistic used to test the hypothesis identified in (a)? ##ANSWER2##
c. What is the relevant sampling distribution for this hypothesis test?  

##ANSWER3##

c. What is the $p$-value computed for the hypothesis test? ##ANSWER4##
d. What is the conclusion from the hypothesis test, at a significance level of `r alpha`.

##ANSWER5##

Question ID: *`r id`*

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(c(options_a[sort_a], options_c[sort_c], options_e[sort_e]), markup = "markdown")
```


Meta-information
================
extype: cloze
exsolution: `r mchoice2string(ans1)`|`r fmt(ans2, 3)`|`r mchoice2string(ans3)`|`r fmt(ans4, 3)`|`r mchoice2string(ans5)`
exclozetype: schoice|num|schoice|num|schoice
exname: TwoSamplePooledVariances
extol: 0.01